### 2024-11-17 main.tex
Intro section

% section/paper on sampling with sampled priors.
%\documentclass[preprint,linenumbers]{aastex63}
\documentclass[linenumbers, onecolumn]{aastex631}
\usepackage{xcolor}

\usepackage{natbib}
\usepackage{amsmath}
\usepackage{enumitem}

\shortauthors{Bernstein et al.}

\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\eqq}[1]{Equation~(\ref{#1})}
\newcommand\gary[1]{{\color{red} \{\textbf{GMB}: #1\}}}
%\tabletypesize{\footnotesize}
\newcommand{\vecc}{\ensuremath{\mathbf{c}}}
\newcommand{\vecq}{\ensuremath{\mathbf{q}}}
\newcommand{\vecn}{\ensuremath{\mathbf{n}}}
\newcommand{\vecu}{\ensuremath{\mathbf{u}}}
\newcommand{\hatc}{\ensuremath{\hat{\mathbf{c}}}}
\newcommand{\vecP}{\ensuremath{\mathbf{P}}}
\newcommand{\likeli}{\mathcal{L}}

\begin{document}

\title{Sampling Bayesian probabilities given only sampled priors}

\author[0000-0002-8613-8259]{Gary M. Bernstein}
\affiliation{Department of Physics and Astronomy, University of Pennsylvania, Philadelphia, PA 19104, USA}
\email{garyb@upenn.edu}

\author{Troxel,\ldots}

\begin{abstract}
	\vspace{0.2in}
Later.
\end{abstract}
\reportnum{}

\section{Introduction} \label{sec:intro}

Our motivating situation is that we have a vector of observable summary statistics \vecc, such as the binned 2-point correlation functions of cosmic fields, that we are using to constrain a set of parameters of interest \vecq, which are cosmological parameters such as $\Omega_m, \sigma_8,$ etc.  There is a model $\vecc(\vecq,\vecn)$ for the observables which involves the parameters of interest, but also a vector \vecn\ of nuisance parameters, which include the coefficients of some linear expansion of the redshift distributions $n(z)$ of some population of the galaxies being observed:
\begin{equation}
  n(z) = \sum_{k=1}^{M} n_k b_k(z).
  \label{eq:nzbasis}
\end{equation}
The $b_k$ are a set of predetermined basis functions for the redshift distribution.  In our case there are multiple galaxy populations to be characterized, leading to hundreds of parameters $n_k$ to be considered.

We wish to characterized the Bayesian posterior probability
\begin{equation}
  p(\vecq | \vecc) \propto \int dn\, \likeli(\vecc | \vecq, \vecn) p(\vecq) p(\vecn),
\label{eq:posterior}
\end{equation}
where $\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function of the data, and $p(\vecq)$ and $p(\vecn)$ are priors on the parameters.  This posterior is complex enough that it requires approximation by the output of a Markov chain wandering across the space $(\vecq,\vecn).$

The scenario of interest is when \emph{the prior $p(\vecn)$ is itself known only from a set of samples of $\vecn$ from this distribution.} Most MC samplers require that the posterior be an evaluable function of any value of the parameters, and it is the general task of density estimators to convert the samples of $\vecn$ into an evaluable $p(\vecn).$  But when \vecn\ is of high dimension, two problems arise: first, there may be insufficient available samples to create a viable density estimator; second, sampling of the posterior in (\ref{eq:posterior}) becomes infeasible if the MC must traverse a high-dimensional space.

One approach would be to run a new MC chain over \vecq\ for each of the samples we have of \vecn, and then concatenate these to effect marginalization over \vecn.  This is clearly infeasible if a large number of \vecn\ samples are needed to characterize the prior in this space.

For the DES Y3 analyses, \citet{hyperrank} devised a scheme whereby the samples of \vecn\ are mapped into a small vector \vecu\ of summary statistics, such as their $\langle z \rangle,$  The \vecu\ space was effectively transformed into a new space $\mathcal{H}$ which is partitioned into regions of equal volume, each containing one sample.  The MC for the cosmology posterior samples over the continuous coordinates of $\mathcal{H}$ with a uniform prior, and the value of \vecn\ that ``owns'' the sampled point of $\mathcal{H}$ space is used to calculate to posterior. This method succeeds in defining a lower-dimensional nuisance space with simple (uniform) prior that yields equal probability to each of the original \vecn\ samples.  The result is, however, discontinuous in the $\mathcal{H}$-space parameters of the nuisance functions, which leads to very inefficient sampling of the posterior.  In particular, samplers such as \textsc{MultiNest} that assume continuity are rendered nearly non-functional.  As a result, the Y3 cosmological priors could not be evaluated with this method.  Instead, the \vecn\ samples were not used, and an \textit{ad hoc} $p(\vecn)$ was adopted which allowed only shifts and dilations of the mean $n(z)$ of the \vecn\ samples.

A more rigorous and extremely efficient method of marginalizing over high-dimensional nuisance parameters was proposed by \citet{hans}, for the case where the following restrictions apply:
\begin{enumerate}
\item The likelihood of the observable \vecc\ is normal, $\vecc \sim \mathcal{N}( \hatc, \Sigma_c).$
\item The prior $p(\vecn)$ can also be assumed to be normal, with a mean and covariance matrix $\Sigma_n$ that in our case could be assigned from the mean and covariance of the samples of \vecn\ we are given.
\item The model \hatc\ can be linearized about fiducial values $\vecq_0, \vecn_0$ without loss of accuracy exceeding measurement errors.
\end{enumerate}
Under these conditions, \citet{hans} show that the marginalization over \vecn\ is equivalent to adding terms to $\Sigma_c,$ such that any MC process need not sample \vecn\ at all.

We describe here an approach that is algebraically similar to that of \citet{hans}, but does not require 2nd condition of Gaussianity for the nuisance prior, and is somewhat more robust to non-linearities in the model that violate the 3rd condition.  Our approach is to seek a linear compression of \vecn\ into a lower-dimensional set of parameters \vecu\ that project away variations in \vecn\ that do not influence the likelihood $\likeli.$  Standard density estimators can then be applied to the \vecu\ values implied by the known \vecn\ samples to yield a prior $p(\vecu)$ that can be used for the MC chain of the cosmological posterior.  The model \hatc, and hence $\likeli,$ will be continuous over this low-dimensional \vecu\ space, and marginalization over \vecu\ will yield posterior probabilities very close to marginalization over the original \vecn.

\begin{acknowledgments}

G.M.B. acknowledges support from NSF grant AST-2205808 and \ldots.

\end{acknowledgments}
% \bibliography{references}
\bibliographystyle{aasjournal}
\bibitem[Cordero et al.(2022)]{hyperrank} Cordero, J.~P., Harrison, I., Rollins, R.~P., et al.\ 2022, \mnras, 511, 2170. doi:10.1093/mnras/stac147

\bibitem[Hadzhiyska et al.(2020)]{hans} Hadzhiyska, B., Alonso, D., Nicola, A., et al.\ 2020, \jcap, 2020, 056. doi:10.1088/1475-7516/2020/10/056

\end{document}

### 2024-11-19 main.tex
typo

@@
 % section/paper on sampling with sampled priors.
-%\documentclass[preprint,linenumbers]{aastex63}
-\documentclass[linenumbers, onecolumn]{aastex631}
+% \documentclass[preprint,linenumbers]{aastex631} # 6.31 fails with \align!!
+\documentclass[linenumbers, onecolumn]{aastex63}
 \usepackage{xcolor}
@@
 \newcommand{\vecP}{\ensuremath{\mathbf{P}}}
+\newcommand{\covm}{C}
+\newcommand{\matA}{\ensuremath{A}}
+\newcommand{\matD}{D}
+\newcommand{\matU}{U}
+\newcommand{\matV}{V}
+\newcommand{\matW}{W}
+\newcommand{\proj}{P}  % Projection matrix
+\newcommand{\jac}{J}   % Jacobian
+\newcommand{\ident}{I}  % identity
+\newcommand{\DD}{\Delta_D}
 \newcommand{\likeli}{\mathcal{L}}
-
+\newcommand{\trace}{\text{Tr}}
 \begin{document}
@@
 
-\section{Introduction} \label{sec:intro}
+\section{Motivation} \label{sec:intro}
 
@@
 \begin{enumerate}
-\item The likelihood of the observable \vecc\ is normal, $\vecc \sim \mathcal{N}( \hatc, \Sigma_c).$
-\item The prior $p(\vecn)$ can also be assumed to be normal, with a mean and covariance matrix $\Sigma_n$ that in our case could be assigned from the mean and covariance of the samples of \vecn\ we are given.
-\item The model \hatc\ can be linearized about fiducial values $\vecq_0, \vecn_0$ without loss of accuracy exceeding measurement errors.
+\item The likelihood of the observable \vecc\ is normal, $\vecc \sim \mathcal{N}( \hatc, \covm_c),$ with $\covm_c$ fixed.
+\item The prior $p(\vecn)$ can also be assumed to be normal, with a mean and covariance matrix $\covm_n$ that in our case could be assigned from the mean and covariance of the samples of \vecn\ we are given.
+\item The model \hatc\ can be linearized in \vecn\ about fiducial values $\vecq_0, \vecn_0$ without loss of accuracy exceeding measurement errors, with the derivatives independent of \vecq.
 \end{enumerate}
-Under these conditions, \citet{hans} show that the marginalization over \vecn\ is equivalent to adding terms to $\Sigma_c,$ such that any MC process need not sample \vecn\ at all.
+Under these conditions, \citet{hans} show that the marginalization over \vecn\ is equivalent to adding terms to $\covm_c,$ such that any MC process need not sample \vecn\ at all.
 
@@
 The model \hatc, and hence $\likeli,$ will be continuous over this low-dimensional \vecu\ space, and marginalization over \vecu\ will yield posterior probabilities very close to marginalization over the original \vecn.
+
+\section{Derivation}\label{sec:deriv}
+We assume that we do have a multivariate normal likelihood for the
+observables \vecc\, with the mean being some model
+$\hatc(\vecq,\vecn)$ and a fixed covariance matrix $\covm_c.$ In this
+section we will assume that the \vecn\ vectors have been shifted by
+the mean $\vecn_0 \equiv \left\langle \vecn_\alpha \right\rangle$ so
+that their mean is zero.
+We seek some function $\hat{\vecn}(\vecu)$ of a lower-dimensional vector \vecu\ which can be substituted for \vecn\ and yield nearly the same likelihood function for any \vecn\ in the domain spanned by the samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha,$ that we are provided.
+This means we want a map $\vecn_\alpha\rightarrow \vecu_\alpha$ and a function $\hat{\vecn}(\vecu)$ which minimize the error in log-likelihood
+\begin{equation} -2 \left\langle \log \likeli(\vecc | \vecq,\vecn) - \log \likeli(\vecc | \vecq,\hat{\vecn}(\vecu)) \right\rangle =
+  \left\langle \Delta\chi^2 \right\rangle \approx \frac{1}{N_\alpha} \sum_\alpha
+                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}(\vecu_\alpha)) \right]^T
+                                            \covm_c^{-1}
+                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}(\vecu_\alpha)) \right].
+                                          \end{equation}
+
+If we next define the Jacobian matrix
+\begin{equation}
+  \jac \equiv \left.\frac{\partial\hatc}{\partial\vecn}\right|_{\vecq_0, \vecn_0}
+\label{eq:jacobian}
+\end{equation}
+and assume that a first-order Taylor expansion about the nominal values is valid across the domain of interest, then we obtain
+\begin{equation}
+  \left\langle \Delta\chi^2 \right\rangle \approx \frac{1}{N_\alpha} \sum_\alpha
+  \left[ \vecn_\alpha - \hat{\vecn}(u_\alpha)\right]^T \jac^T \covm_c^{-1} \jac
+  \left[ \vecn_\alpha - \hat{\vecn}(u_\alpha)\right].
+  \label{eq:linearized}
+\end{equation}
+We ask what is the optimal \emph{linear} model $\hat{\vecn} = \matU \cdot \vecu$ to use.
+It is useful to define the matrix \matD\ and its ``square root'' $\DD$ by
+\begin{equation}
+   \matD \equiv \jac^T \covm_c^{-1} \jac =  \DD^T \DD.
+ \end{equation}
+Since \matD\ is positive-definite and symmetric, $\DD$ will exist and be real-valued.
+If \vecn\ has $N$ elements, and we want to compress the representation
+to a vector \vecu\ with $M$ elements, we will use the $M\times N$ projection
+matrix $\proj_M$ whose elements are
+\begin{equation}
+  \left(\proj_M\right)_{ij} \equiv
+\begin{cases}
+                                            1,  &  i=j\le M \\
+                                            0,  & \text{otherwise.}
+\end{cases}
+\end{equation}
+such that $\proj_M \proj_M^T=\ident_M,$ the $M\times M$ identity
+matrix. We will assume that \matU\ satisfies
+\begin{equation}
+  \DD \matU = \matV_U \proj_M^T
+\label{eq:vupm}
+\end{equation}
+for some unitary matrix $\matV_U$.  We can do this because, for any $\matU$, we have the
+singular-value decomposition $\DD \matU = \matV_U P_M^T \Lambda_U
+W_U^T,$ where $\Lambda_U$ is diagonal, $W_U$ is unitary, and both are
+$M\times M$.  We can always absorb these last two factors into the \vecu\ values
+to conserve the values of $\hat{\vecn} = \matU \vecu$ while making
+\eqq{eq:vupm} hold.
+
+With these definitions and a choice of \matU, the $\chi^2$ error and the values of $\vecu_\alpha$ that minimize it become
+\begin{align}
+  \left\langle \Delta\chi^2 \right\rangle & \approx \frac{1}{N_\alpha} \sum_\alpha
+                                            \left[ \vecn_\alpha - \matU \vecu_\alpha\right]^T \matD^{-1}
+                                            \left[ \vecn_\alpha - \matU \vecu_\alpha\right] \\
+  \Rightarrow \quad \vecu_{\alpha,\text{min}} & = \left[ \matD -
+                                                \matD\matU\left(\matU^{T}\matD\matU\right)^{-1}
+                                                \matU^T\matD\right]\vecn_\alpha  \\
+                                          & = \proj_M \matV_U^T \DD \vecn_\alpha.
+\end{align}                                            
+Defining the covariance matrix of the samples $\vecn_\alpha$ as
+$\covm_n=\left\langle \vecn_\alpha \vecn^T_\alpha \right\rangle,$ and
+plugging the last line back into the first,  gives us the quantity we
+want to minimize by selecting \matU\ properly:
+\begin{align}
+  \left\langle \Delta \chi^2 \right\rangle 
+  & = \trace\left[\covm_n\matD
+    - \covm_n \DD^T \matV_U \proj_M^T \proj_M \matV_U^T \DD \right], \\
+ & = \trace\left[\matV_A \Lambda_A \matV_A^T 
+    - \Lambda_A  \matV_A^T \matV_U \proj_M^T \proj_M \matV_U^T
+   \matV_A\right], \\
+  & = \sum_{i=1}^N \lambda^A_i \left[ 1 - \sum_{j=1}^M (\matV^T_A
+    \matV_U)^2_{ij}\right], \label{eq:dchi2} \\
+  \text{with} \quad \matA & \equiv \DD \covm_c \DD^T = \matV_A \Lambda_A \matV_A^T.
+\end{align}
+The last line defines a new symmetric, non-negative matrix \matA\ that
+has eigenvectors $\matV_A$ and positive eigenvalues $\Lambda_A=\text{Diag}(\lambda^A_1,
+\lambda^A_2, \ldots, \lambda^A_N).$ We assume the eigenvalues are
+placed into decreasing order.
+
+The minimization of $\left\langle \chi^2 \right\rangle$ thus requires
+choosing $\matV_U,$ to minimize \eqq{eq:dchi2}, then retrieving \matU\
+from \eqq{eq:vupm}.  Since $\matV_A^T\matV_U$ is a unitary matrix,
+each term in brackets must be non-negative and $\le 1,$ and their total must equal
+$N-M$.  The optimal case would be to have the bracketed term be zero
+for $i\le M$, and be unity when multiplying the smaller $\lambda^A_i$
+at $i>M.$  This occurs when $\matV_U=\matV_A,$ which is therefore our
+optimal compression in terms of reproducing the likelihood of the
+data.
+
+The desired linear compression of \vecn\ therefore has the following properties:
+\begin{itemize}
+\item The mean error in $2\log \likeli$ is
+  \begin{equation}
+    \left\langle \Delta\chi^2 \right \rangle = \sum_{i=M+1}^N
+    \lambda^A_i,
+  \end{equation}
+  and we will want to choose the dimension $M$ of the compression to
+  keep this quantity, the observable effect of the parts of \vecn\ that we 
+  project away,  below unity.
+  Note that the $\lambda^A_i$ values are determined by a combination of
+  the covariance matrix of the samples of \vecn, along with the Jacobian
+  matrix and the covariance $\covm_c$ of the observables, letting us
+  know the measurement significance of each mode of variation of \vecn.
+
+  \item The \matU\ matrix containing the $M$ basis vectors for
+    reconstructing \vecn\ is
+    \begin{equation}
+      \matU = \DD^{-1} \matV_A \proj_M^T.
+    \end{equation}
+    This requires that $\DD$ be invertible, but if we recall that
+    $\matV_A$ are the eigenvectors of $\DD \covm_n \DD,$ we can see
+    that we can get a well-defined result as long as the rank of
+    \matD\ is $\ge M$.  ???
+
+  \item The compressed $\vecu_\alpha$ for a chosen sample of \vecn\ is
+    \begin{equation}
+      \vecu_\alpha = \proj_M \matV_A^T \DD \vecn_\alpha,
+    \end{equation}
+    and the covariance matrix of the compressed samples is
+    \begin{equation}
+      \covm_u = \left\langle \vecu \vecu^T \right\rangle = \proj_M
+      \matV_A^T \DD \covm_n \DD^T \proj_M^T = \text{Diag}(\lambda^A_1,
+      \lambda^A_2, \ldots \lambda^A_M).
+    \end{equation}
+The components of the compressed \vecu\ coefficients are therefore
+uncorrelated.  Component $u_i$ has a variance
+$\lambda^A_i,$ which is also equal to the reduction in $\left\langle
+  \Delta \chi^2 \right\rangle$ that we will accrue if we include basis
+vector $\mathbf{U}_i$ into our linear reconstruction of \vecn.
+\end{itemize}
+
+Upon executing the compression $\{\vecn_\alpha\} \rightarrow
+\{\vecu_\alpha\},$ we may find the distribution of the samples to be
+consistent with a multivariate normal---which will have the diagonal covariance
+matrix $\covm_u,$ and statistical independence for the elements of
+\vecu.
+There is,however,  no \textit{a prior} reason that
+this must be the case.  We can use any density estimator over
+$\{\vecu_\alpha\}$ to define a continuous $p(\vecu)$ for a Markov
+chain running over the $(\vecq,\vecu)$ space.  This is the principle
+advantage of our method over the single-step covariance-inflation
+method of \citet{hans}.
+
+Even if the \vecu\ prior is well approximated as multivariate normal,
+there are practical advantages of compressing the nuisance 
+variables and retaining them in the cosmological Markov chain rather
+than using the covariance-inflation method for analytic
+marginalization.  One can, for example, examine the posterior
+distributions of \vecu\ and see if they are at the edges of the prior,
+which would potentially indicate an inconsistency between the data and
+the prior.
+
+\section{Application}\label{sec:app}
+
+xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
+xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
+xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
+xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx
 
@@
 \end{acknowledgments}
-% \bibliography{references}
-\bibliographystyle{aasjournal}
+\begin{thebibliography}{dummy}
+%\bibliography
+%\bibliographystyle{aasjournal}
 \bibitem[Cordero et al.(2022)]{hyperrank} Cordero, J.~P., Harrison, I., Rollins, R.~P., et al.\ 2022, \mnras, 511, 2170. doi:10.1093/mnras/stac147
@@
 \bibitem[Hadzhiyska et al.(2020)]{hans} Hadzhiyska, B., Alonso, D., Nicola, A., et al.\ 2020, \jcap, 2020, 056. doi:10.1088/1475-7516/2020/10/056
+\end{thebibliography}
 

### 2024-11-20 main.tex
Fix the inversion issue, add code

@@
 We ask what is the optimal \emph{linear} model $\hat{\vecn} = \matU \cdot \vecu$ to use.
-It is useful to define the matrix \matD\ and its ``square root'' $\DD$ by
+It is useful to define the matrix \matD, its eigensystem
+$(\matV_D,\Lambda_D),$  \ and its ``square root'' $\DD$ by
 \begin{equation}
-   \matD \equiv \jac^T \covm_c^{-1} \jac =  \DD^T \DD.
+   \matD \equiv \jac^T \covm_c^{-1} \jac =  \matV_D \Lambda_D
+   \matV_D^T = \left(\matV_D \Lambda_D^{1/2}\right)
+   \left(\Lambda_D^{1/2}\matV_D^T \right) = \DD^T \DD.
  \end{equation}
-Since \matD\ is positive-definite and symmetric, $\DD$ will exist and be real-valued.
+Since \matD\ is non-negative and symmetric, $\DD$ will exist and be
+real-valued with non-negative singular values.
 If \vecn\ has $N$ elements, and we want to compress the representation
@@
     \begin{equation}
-      \matU = \DD^{-1} \matV_A \proj_M^T.
+      \matU = \DD^{-1} \matV_A \proj_M^T = \matV_D \Lambda_D^{-1/2}  \matV_A \proj_M^T.
     \end{equation}
-    This requires that $\DD$ be invertible, but if we recall that
-    $\matV_A$ are the eigenvectors of $\DD \covm_n \DD,$ we can see
-    that we can get a well-defined result as long as the rank of
-    \matD\ is $\ge M$.  ???
+    In inverting the (non-negative, diagonal) matrix $\Lambda_D^{1/2},$ one should null the
+    inverses of elements that are $\lesssim 10^{-3}$ of the largest
+    element, to avoid adding large fluctuations to \matU\ that address
+    noise or roundoff-error terms that contribute negligibly to $\chi^2.$
 
@@
     \begin{equation}
-      \vecu_\alpha = \proj_M \matV_A^T \DD \vecn_\alpha,
+      \vecu_\alpha = \proj_M \matV_A^T \DD \vecn_\alpha = \proj_M
+      \matV_A^T \Lambda_D^{1/2} \matV_D^T \vecn_\alpha,
     \end{equation}
@@
       \covm_u = \left\langle \vecu \vecu^T \right\rangle = \proj_M
-      \matV_A^T \DD \covm_n \DD^T \proj_M^T = \text{Diag}(\lambda^A_1,
+      \matV_A^T \DD \covm_n \DD^T \matV_A \proj_M^T = \text{Diag}(\lambda^A_1,
       \lambda^A_2, \ldots \lambda^A_M).
@@
 \end{itemize}
+The compression thus consists of choosing the desired number of rows
+of the matrix $ \matV_A^T \Lambda_D^{1/2} \matV_D^T$, and the modes
+are the corresponding columns of this matrix's inverse, with each
+included mode reducing the mean compression error in mean $\chi^2$ by $\lambda^A_i.$
 

### 2024-11-20 nzmodes.py

@@ -1,0 +1,101 @@
+# Codes for creating and smoothing modes
+import numpy as np
+def getModes(D, Cn, nModes, threshold=0.001):
+    '''Calculate the compression matrix from n to u and the
+    modes U that multiply each U.  
+    Parameters:
+    `D`:      The matrix J^T C_c^{-1} J, where C_c is the covariance
+              matrix of the observables c, and J is the Jacobian dc/dn,
+              shape=(N,N)
+    `Cn`:     Covariance matrix of the n(z) parameter vector n, shape=(N,N)
+    `nModes`: Number of modes to retain.
+    `threshold`:  The ratio of smallest singular value retained to the 
+              largest singular value of an internal matrix during a
+              matrix inversion.
+    Returns:
+    `X`:      The compression matrix, u = X @ n, shape=(nModes,N)
+    `U`:      The decoding matrix, n' = U @ u, shape=(N,nModes)
+    `dchidq`: The mean chisq shift generated by each mode, shape=(nModes)
+
+    You can safely use fewer than `nModes` of the returned X and U if you 
+    think the discarded `dchisq` values are small enough.'''
+
+    Dval, Dvec = np.linalg.eigh(D)
+    DeltaD = np.diag(np.sqrt(Dval)) @ Dvec.T # Transposing numpy's eigenvectors
+
+    A = DeltaD @ Cn @ DeltaD.T
+    Aval, Avec = np.linalg.eigh(A)
+    # Sort eigenvalues to be decreasing
+    order = np.argsort(-Aval)
+    Aval = Aval[order]
+    Avec = Avec[:,order]
+
+    X = Avec.T[:nModes] @ DeltaD   # The encoder from n->u
+
+    # This is the SVD of X given how we made it
+    uX = Avec.T
+    sX = np.sqrt(Dval)
+    vX = Dvec
+    
+    # Get U from the SVD-controlled inverse of X
+    thresh = 0.001
+    U = vX @ np.diag(np.where( sX > threshold*np.max(sX), 1/sX, 0.)) @ uX.T[:,:nModes]
+    
+    return X, U, Aval[:nModes]
+
+class Tz:
+    def __init__(self, dz, nz, z0=None):
+        '''Class representing sawtooth n(z) kernels (bins) in z.
+        First kernel is centered at z0, which defaults to dz if not
+        given.  If z0<dz, then its triangle is adjusted to go to zero at
+        0, then peak at z0, down to zero at z0+dz
+        Arguments:
+        `dz`: the step between kernel centers
[…]

### 2024-11-20 main.tex
Better hyperrank description

@@
 
-For the DES Y3 analyses, \citet{hyperrank} devised a scheme whereby the samples of \vecn\ are mapped into a small vector \vecu\ of summary statistics, such as their $\langle z \rangle,$  The \vecu\ space was effectively transformed into a new space $\mathcal{H}$ which is partitioned into regions of equal volume, each containing one sample.
-The MC for the cosmology posterior samples over the continuous coordinates of $\mathcal{H}$ with a uniform prior, and the value of \vecn\ that ``owns'' the sampled point of $\mathcal{H}$ space is used to calculate to posterior.
-This method succeeds in defining a lower-dimensional nuisance space with simple (uniform) prior that yields equal probability to each of the original \vecn\ samples.
-The result is, however, discontinuous in the $\mathcal{H}$-space parameters of the nuisance functions, which leads to very inefficient sampling of the posterior.
+For the DES Y3 analyses, \citet{hyperrank} devised a scheme whereby
+the samples of \vecn\ are placed in a grid within some $M$-dimensional
+unit hypercube $\mathcal{H}$.  The coordinates \vecu\ within the hypercube
+are considered the compressed parameters of $n(z),$ and the
+decompression function
+$\hat{\vecn}(\vecu)$ outputs the $\vecn_\alpha$ sample at the nearest grid point to
+  \vecu.  This solves the problem of creating a continuous \vecu\
+  domain, and maintains the equal prior probability of each \vecn\
+  sample, but the function \emph{output,} and the resultant likelihood
+  function of \vecu\, are discontinuous.
+Various strategies are proposed to assign the $\vecn_\alpha$ to the grid points in
+$\mathcal{H}$ based on summary statistics, to reduce the
+discontinuities.  But the function is never smooth.
+As a consequence, many MC samplers become quite inefficient in
+sampling of the cosmological posterior.
 In particular, samplers such as \textsc{MultiNest} that assume continuity are rendered nearly non-functional.

### 2024-12-13 nzmodes.py
A few typos fixed in latex and a whole new version of python routine.

@@ -1,6 +1,6 @@
 # Codes for creating and smoothing modes
 import numpy as np
-def getModes(D, Cn, nModes, threshold=0.001):
+def getModes(D, Cn, chisq_threshold=0.1):
     '''Calculate the compression matrix from n to u and the
     modes U that multiply each U.  
     Parameters:
@@ -8,40 +8,53 @@ def getModes(D, Cn, nModes, threshold=0.001):
               matrix of the observables c, and J is the Jacobian dc/dn,
               shape=(N,N)
     `Cn`:     Covariance matrix of the n(z) parameter vector n, shape=(N,N)
-    `nModes`: Number of modes to retain.
-    `threshold`:  The ratio of smallest singular value retained to the 
-              largest singular value of an internal matrix during a
-              matrix inversion.
+    `chisq_threshold`:  Your desired upper bound for the mean error in chisq
+              resulting from the compression.
     Returns:
-    `X`:      The compression matrix, u = X @ n, shape=(nModes,N)
+    `X`:      The compression matrix, u = X @ n, shape=(nModes,N).  The
+              covariance matrix of the u's will be the identity.
     `U`:      The decoding matrix, n' = U @ u, shape=(N,nModes)
-    `dchidq`: The mean chisq shift generated by each mode, shape=(nModes)
+    `dchisq`: The mean chisq shift generated by each mode, shape=(nModes)
+    `resid`:  Mean amount of chisq shift caused by compression.
 
     You can safely use fewer than `nModes` of the returned X and U if you 
     think the discarded `dchisq` values are small enough.'''
 
+    thr = 1e-25     # Dynamic range threshold for eigenvalues
+
+    # Symmetrize and sqrt D
+    D = 0.5*(D + D.T)
     Dval, Dvec = np.linalg.eigh(D)
-    DeltaD = np.diag(np.sqrt(Dval)) @ Dvec.T # Transposing numpy's eigenvectors
+    # Any negatives are numerical problems
+    Dval = np.maximum(0., Dval)
 
-    A = DeltaD @ Cn @ DeltaD.T
-    Aval, Avec = np.linalg.eigh(A)
-    # Sort eigenvalues to be decreasing
[…]

### 2024-12-17 main.tex
A few typos fixed in latex and a whole new version of python routine.
Add Modes class to code and put new derivation into latex file

@@
 \newcommand{\vecu}{\ensuremath{\mathbf{u}}}
+\newcommand{\vecv}{\ensuremath{\mathbf{v}}}
 \newcommand{\hatc}{\ensuremath{\hat{\mathbf{c}}}}
@@
 \newcommand{\matD}{D}
-\newcommand{\matU}{U}
+\newcommand{\matE}{E}
+\newcommand{\matF}{F}
+\newcommand{\matG}{G}
+\newcommand{\matI}{I}
+\newcommand{\matT}{T}
+\newcommand{\matX}{X}
+\newcommand{\matY}{Y}
 \newcommand{\matV}{V}
-\newcommand{\matW}{W}
+\newcommand{\matLam}{\Lambda}
 \newcommand{\proj}{P}  % Projection matrix
@@
 
-We describe here an approach that is algebraically similar to that of \citet{hans}, but does not require 2nd condition of Gaussianity for the nuisance prior, and is somewhat more robust to non-linearities in the model that violate the 3rd condition.
+We describe here an approach that is algebraically similar to that of \citet{hans}, but does not require 2nd condition of Gaussianity for the nuisance prior.
 Our approach is to seek a linear compression of \vecn\ into a lower-dimensional set of parameters \vecu\ that project away variations in \vecn\ that do not influence the likelihood $\likeli.$  Standard density estimators can then be applied to the \vecu\ values implied by the known \vecn\ samples to yield a prior $p(\vecu)$ that can be used for the MC chain of the cosmological posterior.
@@
 that their mean is zero.
-We seek some function $\hat{\vecn}(\vecu)$ of a lower-dimensional vector \vecu\ which can be substituted for \vecn\ and yield nearly the same likelihood function for any \vecn\ in the domain spanned by the samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha,$ that we are provided.
-This means we want a map $\vecn_\alpha\rightarrow \vecu_\alpha$ and a function $\hat{\vecn}(\vecu)$ which minimize the error in log-likelihood
-\begin{equation} -2 \left\langle \log \likeli(\vecc | \vecq,\vecn) - \log \likeli(\vecc | \vecq,\hat{\vecn}(\vecu)) \right\rangle =
-  \left\langle \Delta\chi^2 \right\rangle \approx \frac{1}{N_\alpha} \sum_\alpha
-                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}(\vecu_\alpha)) \right]^T
+We seek some function $\hat{\vecn}(\vecu)$ of a lower-dimensional
+vector \vecu\ which can be substituted for \vecn\ and yield nearly the
+same likelihood function for any \vecn\ in the domain spanned by the
+samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha,$ that we are
+provided.  This means we want a map $\vecn_\alpha\rightarrow
+\vecu_\alpha \rightarrow \hat{\vecn}_\alpha$ such that replacing $\vecn$ with
+$\hat\vecn$ alters the cosmological inference by much less than the
+other uncertainties.  We will implement this by minimizing the
+distance in the space $\vecc$ between the
+model generated by $\vecn$ and that by $\hat\vecn$, using the
+observations' covariance matrix $\covm_c$ as a metric for the
+distance.  This is equivalent to the $\chi^2$ of the difference
+between the original and compressed models for the data:
+\begin{equation} \left\langle \chi^2 \right\rangle
+=  \frac{1}{N_\alpha} \sum_\alpha
+                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}_\alpha) \right]^T
                                             \covm_c^{-1}
-                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}(\vecu_\alpha)) \right].
+                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}\alpha) \right].
+\label{eq:chisq}
 \end{equation}
+If the data are in fact drawn from the model $\hatc(\vecq,\vecn)$ with
+a Gaussian likelihood, then this is also the mean shift in $-2\log\likeli$
+from the compression.  It is \emph{not}, however, equal to the mean
+shift of the overall log of the posterior in
+\eqq{eq:posterior}---$\left\langle\chi^2\right\rangle$ is serving as a
+proxy for this.
 
-If we next define the Jacobian matrix
+We next assume that the compression is linear, $\hat\vecn=\matX\vecn,$
+for some matrix that is idempotent ($\matX\matX = \matX$).  If this is
+true, then the contribution to \eqq{eq:chisq} at lowest order in
+$\vecn-\hat\vecn$ is
 \begin{equation}
-  \jac \equiv \left.\frac{\partial\hatc}{\partial\vecn}\right|_{\vecq_0, \vecn_0}
-\label{eq:jacobian}
-\end{equation}
-and assume that a first-order Taylor expansion about the nominal values is valid across the domain of interest, then we obtain
-\begin{equation}
-  \left\langle \Delta\chi^2 \right\rangle \approx \frac{1}{N_\alpha} \sum_\alpha
-  \left[ \vecn_\alpha - \hat{\vecn}(u_\alpha)\right]^T \jac^T \covm_c^{-1} \jac
-  \left[ \vecn_\alpha - \hat{\vecn}(u_\alpha)\right].
+  \left\langle \chi^2 \right\rangle = \frac{1}{N_\alpha} \sum_\alpha
+  \left[ (\matI-\matX)\vecn_\alpha\right]^T \matF  \left[ (\matI-\matX)\vecn_\alpha\right]
   \label{eq:linearized}
 \end{equation}
-We ask what is the optimal \emph{linear} model $\hat{\vecn} = \matU \cdot \vecu$ to use.
-It is useful to define the matrix \matD, its eigensystem
-$(\matV_D,\Lambda_D),$  \ and its ``square root'' $\DD$ by
+where we use the Jacobian matrix of the model $\hatc$ to define
 \begin{equation}
-   \matD \equiv \jac^T \covm_c^{-1} \jac =  \matV_D \Lambda_D
-   \matV_D^T = \left(\matV_D \Lambda_D^{1/2}\right)
-   \left(\Lambda_D^{1/2}\matV_D^T \right) = \DD^T \DD.
+  \matF \equiv
+  \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0, \vecn_0}
+  \covm_c^{-1} \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0,
+    \vecn_0}^T.
+\label{eq:fisher}
 \end{equation}
-Since \matD\ is non-negative and symmetric, $\DD$ will exist and be
-real-valued with non-negative singular values.
-If \vecn\ has $N$ elements, and we want to compress the representation
-to a vector \vecu\ with $M$ elements, we will use the $M\times N$ projection
-matrix $\proj_M$ whose elements are
+This quantity is also the Fisher matrix over $\vecn$ giving the information
+provided by the observations $\vecc.$  In many cases this matrix will be
+rank-deficient and/or poorly conditioned, since the observables are not
+likely to be very informative on $\vecn$---if they were, we might not be
+concerned with establishing a prior on $\vecn$ to begin with.  We will
+not, however, require its inverse.
+
+Since $\matX$ is idempotent, we can write
+\begin{align}
+  \matX & = V_X \proj_M V_X^T, \\
+  \matY \equiv \matI-\matX & = V_X \proj_{-M} V_X^T,
+\end{align}
+where $V_X$ is unitary and the projection matrix $\proj_M$ is defined as
 \begin{equation}
@@
                                             1,  &  i=j\le M \\
-                                            0,  & \text{otherwise.}
+                                            0,  & \text{otherwise}
 \end{cases}
 \end{equation}
-such that $\proj_M \proj_M^T=\ident_M,$ the $M\times M$ identity
-matrix. We will assume that \matU\ satisfies
+and $\proj_{-M}=\matI-\proj_M.$  For a chosen rank $M$ of the
+transformation matrix $X$, our task becomes to identify the
+eigenvectors $V_X$ that minimize
+\begin{align}
+  \left\langle \chi^2\right\rangle & = \frac{1}{N_\alpha} \sum_\alpha
+  \left[ \matY \vecn_\alpha\right]^T \matF  \left[\matY
+                                     \vecn_\alpha\right] \\
+       & = \trace \left[ \covm_n V_X \proj_{-M} V_X^T \matF V_X
+         \proj_{-M} V_X^T \right].
+\end{align}
+This optimization is easier if we transform first the systematic variables to
+$\vecn^\prime =\matT\vecn$ such that $\covm_{n^\prime}=\matI$, \ie\
+make the elements of $\vecn$ uncorrelated and unit-variance.  This
+is accomplished by finding the eigensystem $\covm_n=\matV_n \matLam_n
+\matV_n^T$ and setting $\matT = \matLam_n^{-1/2} \matV_n^T$.  With
+this transformation, we are now seeking a different unitary matrix
+$\matV_{X^\prime}$ that minimizes
+\begin{align}
+  \left\langle \chi^2\right\rangle & = \trace \left[ \matI
+    V_{X^\prime} \proj_{-M} V_{X^\prime}^T \left[ (T^{-1})^T \matF
+      T^{-1} \right] V_{X^\prime} 
+    \proj_{-M} V_{X^\prime}^T \right] \\
+  & = \trace \left[ \proj_{-M} V_{X^\prime}^T \matV_G \matLam_G \matV_G^T
+    V_{X^\prime} \proj_{-M} \right],
+  \label{eq:mintr}\\
+  \matG \equiv \left(T^{-1}\right)^T \matF   T^{-1} & = \matLam_n^{1/2} \matV_n^T
+  \matF \matV_n \matLam_n^{1/2} = \matV_G \matLam_G \matV_G^T.
+  \label{eq:defG}
+\end{align}
+The last line defines a transformed Fisher matrix $\matG$ and its
+(non-negative) eigensystem.  Examination of \eqq{eq:mintr} reveals that this quantity
+is minimized if the elements of the unitary matrices satisfy
+$\matV_{X^\prime}^T \matV_G = \matI \; \Rightarrow \; \matV_{X^\prime} = \matV_G,$ and the eigensystem of $G$ is placed
+in order of decreasing eigenvalue $\lambda^G_i.$ The
+elements surviving the project functions yield
 \begin{equation}
-  \DD \matU = \matV_U \proj_M^T
-\label{eq:vupm}
+  \left\langle \chi^2\right\rangle = \sum_{i>M} \lambda_i^G.
+  \label{eq:chiresid}
 \end{equation}
-for some unitary matrix $\matV_U$.  We can do this because, for any $\matU$, we have the
-singular-value decomposition $\DD \matU = \matV_U P_M^T \Lambda_U
-W_U^T,$ where $\Lambda_U$ is diagonal, $W_U$ is unitary, and both are
-$M\times M$.  We can always absorb these last two factors into the \vecu\ values
-to conserve the values of $\hat{\vecn} = \matU \vecu$ while making
-\eqq{eq:vupm} hold.
+In other words each eigenvalue of the matrix $\matG$ in \eqq{eq:defG}
+gives the contribution to $\left\langle\chi^2\right\rangle$ of one
+projection (mode) of $\vecn.$
 
-With these definitions and a choice of \matU, the $\chi^2$ error and the values of $\vecu_\alpha$ that minimize it become
+Transforming the solution back into the space of $\vecn$ yields
 \begin{align}
-  \left\langle \Delta\chi^2 \right\rangle & \approx \frac{1}{N_\alpha} \sum_\alpha
-                                            \left[ \vecn_\alpha - \matU \vecu_\alpha\right]^T \matD^{-1}
-                                            \left[ \vecn_\alpha - \matU \vecu_\alpha\right] \\
-  \Rightarrow \quad \vecu_{\alpha,\text{min}} & = \left[ \matD -
-                                                \matD\matU\left(\matU^{T}\matD\matU\right)^{-1}
-                                                \matU^T\matD\right]\vecn_\alpha  \\
-                                          & = \proj_M \matV_U^T \DD \vecn_\alpha.
+  \matX & = \matT^{-1} \matV_G \proj_M \matV_G^T \matT \\
+   & = \left[ \matV_n^T \matLam_n^{1/2} \matV_G \proj_M \right] \left[
+     \proj_M \matV_G^T \matLam_n^{-1/2} \matV_n^T \right] \\
+\label{eq:DE}
+        & \equiv \matD \matE.
 \end{align}
-Defining the covariance matrix of the samples $\vecn_\alpha$ as
-$\covm_n=\left\langle \vecn_\alpha \vecn^T_\alpha \right\rangle,$ and
-plugging the last line back into the first,  gives us the quantity we
-want to minimize by selecting \matU\ properly:
-\begin{align}
-  \left\langle \Delta \chi^2 \right\rangle 
-  & = \trace\left[\covm_n\matD
-    - \covm_n \DD^T \matV_U \proj_M^T \proj_M \matV_U^T \DD \right], \\
- & = \trace\left[\matV_A \Lambda_A \matV_A^T 
-    - \Lambda_A  \matV_A^T \matV_U \proj_M^T \proj_M \matV_U^T
-   \matV_A\right], \\
-  & = \sum_{i=1}^N \lambda^A_i \left[ 1 - \sum_{j=1}^M (\matV^T_A
-    \matV_U)^2_{ij}\right], \label{eq:dchi2} \\
-  \text{with} \quad \matA & \equiv \DD \covm_c \DD^T = \matV_A \Lambda_A \matV_A^T.
-\end{align}
-The last line defines a new symmetric, non-negative matrix \matA\ that
-has eigenvectors $\matV_A$ and positive eigenvalues $\Lambda_A=\text{Diag}(\lambda^A_1,
-\lambda^A_2, \ldots, \lambda^A_N).$ We assume the eigenvalues are
-placed into decreasing order.
+We thus obtain our optimal encoding/compression using the nonzero rows of matrix
+$\matE$ to give
+\begin{equation}
+  \vecu_\alpha = \matE \vecn_\alpha
+  \label{eq:uu}
+\end{equation}
+and the decoding/reconstruction of the systematic variables as
+\begin{equation}
+  \hat\vecn_\alpha = \matD \vecu_\alpha.
+  \label{eq:UU}
+\end{equation}
+One can confirm that this procedure yields a compressed representation
+$\vecu$ such that $\covm_u = \matI_M,$ the $M$-dimensional
+identity matrix.
 
-The minimization of $\left\langle \chi^2 \right\rangle$ thus requires
-choosing $\matV_U,$ to minimize \eqq{eq:dchi2}, then retrieving \matU\
-from \eqq{eq:vupm}.  Since $\matV_A^T\matV_U$ is a unitary matrix,
-each term in brackets must be non-negative and $\le 1,$ and their total must equal
-$N-M$.  The optimal case would be to have the bracketed term be zero
-for $i\le M$, and be unity when multiplying the smaller $\lambda^A_i$
-at $i>M.$  This occurs when $\matV_U=\matV_A,$ which is therefore our
-optimal compression in terms of reproducing the likelihood of the
-data.
+The previous derivation ignores the possibility that $\covm_n$ is
+singular or nearly so, such that taking $\matLam_n^{-1/2}$ in
+\eqq{eq:DE} is not possible.  Indeed in our application, it is
+\emph{required} that $\covm_n$ be singular, because we have a sum
+normalization constraint on the initial $\vecn_\alpha$ values.  Any
+such (nearly) zero element $j$ of $\matLam_n$ has a corresponding
+eigenvector $\vecv_j$ of the $\vecn$ space which has zero amplitude
+in all of the input samples $\vecn_\alpha,$ so that the $\vecn$ are
+confined to a subspace---the reconstructed $\hat\vecn$ should also be.
+The compressed
+representations $\vecu_\alpha$ and reconstructed $\hat\vecn_\alpha$
+should therefore be unaffected by the presence of any $\vecv_j$
+component.  This can be accomplished in \eqq{eq:DE} by setting element
+$j$ of $\matLam_n^{-1/2}$ to zero, as is typically done during
+solutions of least-squares problems using singular value
+decompositions.
 
-The desired linear compression of \vecn\ therefore has the following properties:
-\begin{itemize}
-\item The mean error in $2\log \likeli$ is
-  \begin{equation}
-    \left\langle \Delta\chi^2 \right \rangle = \sum_{i=M+1}^N
-    \lambda^A_i,
-  \end{equation}
-  and we will want to choose the dimension $M$ of the compression to
-  keep this quantity, the observable effect of the parts of \vecn\ that we 
-  project away,  below unity.
-  Note that the $\lambda^A_i$ values are determined by a combination of
-  the covariance matrix of the samples of \vecn, along with the Jacobian
-  matrix and the covariance $\covm_c$ of the observables, letting us
-  know the measurement significance of each mode of variation of \vecn.
+In summary, the procedure for dimensional reduction is:
+\begin{enumerate}
+  \item Obtain the Fisher matrix $\matF$ of the system defined in
+    \eqq{eq:fisher} using derivatives about the mean sample $\vecn_0,$
+    plus the covariance matrix $\covm_n$ of the samples.
+  \item From the eigensystem $(\matLam_n,\matV_n)$ of $C_n$, form the
+    matrix $G$ defined in \eqq{eq:defG} and get its eigensystem
+    $(\matLam_G,\matV_G).$  Place the eigenvalues in descending order.
+  \item Choose the size $M$ of the compressed representation to be the
+    minimum that keeps the $\left\langle\chi^2\right\rangle$ value in
+    \eqq{eq:chiresid} below a chosen threshold, presumably $\ll 1.$
+  \item The encoding matrix $E$ and decoding matrix $D$ are formed as
+    in \eqq{eq:DE}, taking the inverse $\matLam_n^{-1/2}$ to be zero
+    for any eigenvalues that are zero (or within roundoff errors).
+  \item Compress all incoming (mean-subtracted) samples using \eqq{eq:uu}.  They will
+    have unit covariance matrix and zero mean.
+  \item Construct a continuous density estimator in $\vecu$ space that
+    mimics the finite sample distribution.  If the distribution is
+    normal, this becomes the multidimensional unit normal.  There
+    is, however,  no \textit{a prior} reason that this must be the case,
+    and something like a normalizing flow will be needed to
+    approximate this lower-dimensional representation of the prior.
+  \item Sample over $\vecu$ space in the Markov Chain that is sampling the
+    posterior on the parameters of interest $\vecq,$ using \eqq{eq:UU}
+    to transform each sample back into a $\hat\vecn$ vector.
+  \end{enumerate}
 
-  \item The \matU\ matrix containing the $M$ basis vectors for
-    reconstructing \vecn\ is
-    \begin{equation}
-      \matU = \DD^{-1} \matV_A \proj_M^T = \matV_D \Lambda_D^{-1/2}  \matV_A \proj_M^T.
-    \end{equation}
-    In inverting the (non-negative, diagonal) matrix $\Lambda_D^{1/2},$ one should null the
-    inverses of elements that are $\lesssim 10^{-3}$ of the largest
-    element, to avoid adding large fluctuations to \matU\ that address
-    noise or roundoff-error terms that contribute negligibly to $\chi^2.$
-
-  \item The compressed $\vecu_\alpha$ for a chosen sample of \vecn\ is
-    \begin{equation}
-      \vecu_\alpha = \proj_M \matV_A^T \DD \vecn_\alpha = \proj_M
-      \matV_A^T \Lambda_D^{1/2} \matV_D^T \vecn_\alpha,
-    \end{equation}
-    and the covariance matrix of the compressed samples is
-    \begin{equation}
-      \covm_u = \left\langle \vecu \vecu^T \right\rangle = \proj_M
-      \matV_A^T \DD \covm_n \DD^T \matV_A \proj_M^T = \text{Diag}(\lambda^A_1,
-      \lambda^A_2, \ldots \lambda^A_M).
-    \end{equation}
-The components of the compressed \vecu\ coefficients are therefore
-uncorrelated.  Component $u_i$ has a variance
-$\lambda^A_i,$ which is also equal to the reduction in $\left\langle
-  \Delta \chi^2 \right\rangle$ that we will accrue if we include basis
-vector $\mathbf{U}_i$ into our linear reconstruction of \vecn.
-\end{itemize}
-The compression thus consists of choosing the desired number of rows
-of the matrix $ \matV_A^T \Lambda_D^{1/2} \matV_D^T$, and the modes
-are the corresponding columns of this matrix's inverse, with each
-included mode reducing the mean compression error in mean $\chi^2$ by $\lambda^A_i.$
-
-Upon executing the compression $\{\vecn_\alpha\} \rightarrow
-\{\vecu_\alpha\},$ we may find the distribution of the samples to be
-consistent with a multivariate normal---which will have the diagonal covariance
-matrix $\covm_u,$ and statistical independence for the elements of
-\vecu.
-There is,however,  no \textit{a prior} reason that
-this must be the case.  We can use any density estimator over
-$\{\vecu_\alpha\}$ to define a continuous $p(\vecu)$ for a Markov
-chain running over the $(\vecq,\vecu)$ space.  This is the principle
+The ability to accomodate non-Gaussian distributions of the
+nuisance-parameter space is the principle
 advantage of our method over the single-step covariance-inflation
 method of \citet{hans}.
-
-Even if the \vecu\ prior is well approximated as multivariate normal,
+But even if the \vecu\ prior is well approximated as multivariate normal,
 there are practical advantages of compressing the nuisance 

### 2024-12-17 nzmodes.py

@@ -56,6 +56,24 @@ def getModes(D, Cn, chisq_threshold=0.1):
     
     return X, U, s*s, resid
 
+class Modes:
+    def __init__(self, file):
+        '''Compress or decompress n(z)'s into modes defined in the given file.'''
+        self.a = np.load(file)
+        
+    def compress(self,nz):
+        '''Compress n(z) array into mode coefficients.
+        Argument is an (...,N) array of n(z) values that is already
+        normalized according to conventions used to create modes.
+        
+        Returns (...,M) array of mode coefficients from projection.'''
+        return np.einsum('...j,kj->...k',nz-self.a['mean'],self.a['X'])
+    def decompress(self,u):
+        '''Reconstruct n(z) from coefficients.
+        Argument is (...,M) array of mode coefficients.
+        Returns (...,N) array of n(z) values.'''
+        return np.einsum('...j,kj->...k',u, self.a['U']) + self.a['mean']
+ 
 class Tz:
     def __init__(self, dz, nz, z0=None):
         '''Class representing sawtooth n(z) kernels (bins) in z.
@@ -112,3 +130,5 @@ class Tz:
         # Make the kernel coefficients at the z's
         kk = np.array([self(k,z) for k in range(self.nz)])
         return np.einsum('...i,ij->...j',coeffs,kk) / np.sum(coeffs, axis=-1)
+
+ 

### 2024-12-18 nzmodes.py
Fix Tz bug

@@ -129,6 +129,6 @@ class Tz:
         
         # Make the kernel coefficients at the z's
         kk = np.array([self(k,z) for k in range(self.nz)])
-        return np.einsum('...i,ij->...j',coeffs,kk) / np.sum(coeffs, axis=-1)
+        return np.einsum('...i,ij->...j',coeffs,kk) / np.sum(coeffs, axis=-1)[...,np.newaxis]
 
  

### 2024-12-19 nzmodes.py
Add Normalizer

@@ -1,5 +1,8 @@
 # Codes for creating and smoothing modes
 import numpy as np
+from scipy.stats import norm
+from scipy.interpolate import interp1d
+
 def getModes(D, Cn, chisq_threshold=0.1):
     '''Calculate the compression matrix from n to u and the
     modes U that multiply each U.  
@@ -132,3 +135,32 @@ class Tz:
         return np.einsum('...i,ij->...j',coeffs,kk) / np.sum(coeffs, axis=-1)[...,np.newaxis]
 
  
+class Normalizer:
+    def __init__(self, x, gmax=5, kind='cubic'):
+        '''Class to create transformations from some sampled distribution x 
+        to a new variable g that is distributed as a unit normal.  Supply
+        the x distribution at initialization, then the methods `gauss`
+        and `degauss` will be maps from x->g and g->x, respectively.
+        WARNING: these will not be exact inverses of each other because
+        of the use of spline interpolation.
+        `gmax` gives the bounds in g space of the interpolator, i.e. number
+        of sigma.'''
+        
+        # Establishing matching percentile points
+        dg = 0.1
+        gg = np.arange(-gmax,gmax+dg/2, dg)
+        xx = np.percentile(x, norm.cdf(gg)*100)
+        # Clean out any duplicates in the tables.  This is hacky.
+        dup = xx[1:]==xx[:-1]
+        keep = np.concatenate( (np.logical_not(dup), [True]))
+        gg = gg[keep]
+        xx = xx[keep]
+        self.fwd = interp1d(xx,gg,kind=kind, bounds_error=False, fill_value=(-gmax,gmax))
+        self.inv = interp1d(gg,xx,kind=kind, bounds_error=False, fill_value=(np.min(x),np.max(x)))
+        return
+    def gauss(self,x):
+        '''Return Gaussianized values of x'''
+        return self.fwd(x)
+    def degauss(self,g):
+        '''Return de-Gaussianized values of g'''
+        return self.inv(g)

### 2025-01-20 main.tex
Some proofing and text for application.

@@
 Our motivating situation is that we have a vector of observable summary statistics \vecc, such as the binned 2-point correlation functions of cosmic fields, that we are using to constrain a set of parameters of interest \vecq, which are cosmological parameters such as $\Omega_m, \sigma_8,$ etc.
-There is a model $\vecc(\vecq,\vecn)$ for the observables which involves the parameters of interest, but also a vector \vecn\ of nuisance parameters, which include the coefficients of some linear expansion of the redshift distributions $n(z)$ of some population of the galaxies being observed:
+There is a model $\hatc(\vecq,\vecn)$ for the observables which involves the parameters of interest, but also a vector \vecn\ of nuisance parameters, which include the coefficients of some linear expansion of the redshift distributions $n(z)$ of some population of the galaxies being observed:
 \begin{equation}
@@
 \end{equation}
-where $\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function of the data, and $p(\vecq)$ and $p(\vecn)$ are priors on the parameters.
-This posterior is complex enough that it requires approximation by the output of a Markov chain wandering across the space $(\vecq,\vecn).$
+where $\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function
+of the data, and $p(\vecq)$ and $p(\vecn)$ are priors on the
+parameters.  This posterior is complex enough that it requires
+approximation by the output of a Markov chain (MC) wandering across the space $(\vecq,\vecn).$
 
@@
   sample, but the function \emph{output,} and the resultant likelihood
-  function of \vecu\, are discontinuous.
+  function of \vecu, are discontinuous.
 Various strategies are proposed to assign the $\vecn_\alpha$ to the grid points in
@@
 same likelihood function for any \vecn\ in the domain spanned by the
-samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha,$ that we are
-provided.  This means we want a map $\vecn_\alpha\rightarrow
+samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha.$
+This means we want a map $\vecn_\alpha\rightarrow
 \vecu_\alpha \rightarrow \hat{\vecn}_\alpha$ such that replacing $\vecn$ with
@@
 shift of the overall log of the posterior in
-\eqq{eq:posterior}---$\left\langle\chi^2\right\rangle$ is serving as a
-proxy for this.
+\eqq{eq:posterior}---rather, $\left\langle\chi^2\right\rangle$ is serving as a
+proxy for the true log-likelihood shift.
 
@@
 concerned with establishing a prior on $\vecn$ to begin with.  We will
-not, however, require its inverse.
+not, however, require the inverse of $\matF$ to proceed.
+
+The optimization implied by \eqq{\eq:linearized} is the same as in
+familiar Principal Components Analysis (PCA), aside from the presence
+of the $F$ matrix, which in essence defines a new metric for the
+variance to be captured by the principal components.  Our solution will follow the typical derivation
+for PCA, but with an additional variable transformation to compensate
+for the presence of $\matF.$
 
@@
   \matX & = \matT^{-1} \matV_G \proj_M \matV_G^T \matT \\
+\label{eq:DE}
    & = \left[ \matV_n^T \matLam_n^{1/2} \matV_G \proj_M \right] \left[
      \proj_M \matV_G^T \matLam_n^{-1/2} \matV_n^T \right] \\
-\label{eq:DE}
         & \equiv \matD \matE.
@@
     for any eigenvalues that are zero (or within roundoff errors).
-  \item Compress all incoming (mean-subtracted) samples using \eqq{eq:uu}.  They will
+  \item Compress all incoming (mean-subtracted) samples using
+    \eqq{eq:uu}.  The resulting $\vecu$ values will have
     have unit covariance matrix and zero mean.
@@
     normal, this becomes the multidimensional unit normal.  There
-    is, however,  no \textit{a prior} reason that this must be the case,
-    and something like a normalizing flow will be needed to
+    is, however,  no \textit{a priori} reason that this must be the case,
+    and something like a normalizing flow may be needed to
     approximate this lower-dimensional representation of the prior.
@@
 \section{Application}\label{sec:app}
+As an example of the application of this straightforward dimensional
+reduction to a high-dimensional nuisance parameter, we examine the
+redshift distribution of one of the bins of ``Maglim''  galaxies used
+as a lens population and clustering tracer in the Year 6 analysis of
+the \textit{Dark Energy Survey (DES)} galaxy catalogs.  In the baseline case,
+there are 6 cosmological parameters of interest, $\vecq = \{\Omega_m,
+\Omega_b, \sigma_8, h, n_s, m_\nu\}.$   The nuisance vector $\vecn$
+has 5--10 parameters in each of the following categories: galaxy
+biases with respect to matter; intrinsic alignments of galaxy shapes
+with the tidal field of the mass; and multiplicative errors in the
+measurement of galaxy shear.  The bulk of the nuisance parameters,
+however, are the descriptions of the redshift distribution $n(z)$ of
+galaxies belonging to each of 10 distinct color-based selections
+applied to the galaxy catalog. Each galaxy sample's $n(z)$ distribution is
+described by \eqq{eq:nzbasis} with 80 coefficients spanning $0<z<4$ at
+intervals of $\Delta z=0.05,$  giving a total of 800 nuisance
+parameters.  Each galaxy sample's $n(z)$ is also constrained to
+integrate to unity, which yields 10 linear constraints on these
+nuisance parameters.
 
-xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
-xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
-xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
-xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx
+Various forms of information from spectroscopic and photometric galaxy
+surveys are used to constrain the $n(z)$ coefficients. As detailed in
+\citet{y6pz}, these generate a set of samples of possible $n(z)$
+coefficients for each galaxy sample.  We show here the results of the
+mode compression process as applied to one of the nine galaxy samples,
+the ??th lowest redshift of 6 lens bins.  We wish to have the
+cosmological Markov chain sample from this
+galaxy sample's 80-vector vector $\vecn$ of coefficients for its
+redshift distribution, being guided by the available 3000??
+equal-probability samples.
+
+The observable quantities $\vecc$ whose modeling depends upon the these nuisance
+parameters are the angular autocorrelation $w(\theta)$ of the sample
+members, and the cross-correlations $\gamma_t^{(s)}(\theta)$ between
+this sample's positions and the weak gravitational lensing shear
+observed on the samples $1\le s \le 4$ of source galaxies.  A
+covariance matrix spanning all of the experiment's observables is
+available, and for this demonstration we will restrict it to the 30??
+total elements of $\vecc$ that are affected by the $n(z)$ model for
+this bin.
+
+Figure showing violin plot, before/after.
+
+Figure showing modes
+
+Figure showing $\vecu$ distributions.
+
+Describe 1d normalization mapping.
+
+Figure showing histogram of chisq shifts.
 
@@
 
+\bibiten[TBD]{y6pz} Boyan, William, ??? et al.\ 2025 (in preparation)
+Y6 redshift paper(s).
+
 \end{document}

### 2025-01-31 main.tex
Finished draft

@@
 \newcommand{\vecq}{\ensuremath{\mathbf{q}}}
+\newcommand{\vecs}{\ensuremath{\mathbf{s}}}
 \newcommand{\vecn}{\ensuremath{\mathbf{n}}}
@@
 
-\author{Troxel,\ldots}
+\author{Troxel,William,Boyan,Giulia,Alex Amon,Alex Alarcon,\ldots}
 
@@
 	\vspace{0.2in}
-Later.
+A typical Bayesian inference on the values of some parameters of
+interest $\vecq$ from some data $D$ would run a Markov Chain (MC) to sample
+from the posterior $p(\vecq,\vecn | D) \propto \likeli(D | \vecq,\vecn)
+p(\vecq) p(\vecn),$ where $\vecn$ are some nuisance parameters.  In
+many cases, the nuisance parameters are high-dimensional, and their
+prior $p(\vecn)$ is itself defined only by a set of samples that have
+been drawn from some other MC.  Two problems arise: first,
+the MC for the posterior will typically require evaluation of
+$p(\vecn)$ at arbitrary values of $\vecn,$ \ie\ one needs to provide a
+density estimator over the full $\vecn$ space from the provided
+samples.  Second, the high dimensionality of $\vecn$ hinders both the
+density estimation and the efficiency of the MC for the posterior.  We
+describe a solution to this problem: a linear compression of the
+$\vecn$ space into a much lower-dimensional space $\vecu$ which
+projects away directions in $\vecn$ space that cannot appreciably
+alter $\likeli.$ The algorithm for doing so is a slight modification
+to principal components analysis, and is less restrictive on
+$p(\vecn)$ than other proposed solutions to this issue.
+We demonstrate this ``mode projection'' technique using the analysis
+of 2-point correlation functions of weak lensing fields and galaxy
+density in the \textit{Dark Energy Survey}, where $\vecn$ is a binned representation of the redshift distrbution
+$n(z)$ of the galaxies.
 \end{abstract}
@@
 
-Our motivating situation is that we have a vector of observable summary statistics \vecc, such as the binned 2-point correlation functions of cosmic fields, that we are using to constrain a set of parameters of interest \vecq, which are cosmological parameters such as $\Omega_m, \sigma_8,$ etc.
-There is a model $\hatc(\vecq,\vecn)$ for the observables which involves the parameters of interest, but also a vector \vecn\ of nuisance parameters, which include the coefficients of some linear expansion of the redshift distributions $n(z)$ of some population of the galaxies being observed:
-\begin{equation}
-  n(z) = \sum_{k=1}^{M} n_k b_k(z).
-  \label{eq:nzbasis}
-\end{equation}
-The $b_k$ are a set of predetermined basis functions for the redshift distribution.
-In our case there are multiple galaxy populations to be characterized, leading to hundreds of parameters $n_k$ to be considered.
-
+Consider an inference in which we have a vector of observable summary
+statistics \vecc\ that we are using to constrain a set of parameters of
+interest \vecq.  There is a model $\hatc(\vecq,\vecn)$ for the
+observables which involves the parameters of interest, but also a
+vector \vecn\ of nuisance parameters.  
 We wish to characterized the Bayesian posterior probability
@@
 
-The scenario of interest is when \emph{the prior $p(\vecn)$ is itself known only from a set of samples of $\vecn$ from this distribution.} Most MC samplers require that the posterior be an evaluable function of any value of the parameters, and it is the general task of density estimators to convert the samples of $\vecn$ into an evaluable $p(\vecn).$  But when \vecn\ is of high dimension, two problems arise: first, there may be insufficient available samples to create a viable density estimator; second, sampling of the posterior in (\ref{eq:posterior}) becomes infeasible if the MC must traverse a high-dimensional space.
+The scenario we address here is when \emph{the prior $p(\vecn)$ is itself known only from a set of samples of $\vecn$ from this distribution.} Most MC samplers require that the posterior be an evaluable function of any value of the parameters, and it is the general task of density estimators to convert the samples of $\vecn$ into an evaluable $p(\vecn).$  But when \vecn\ is of high dimension, two problems arise: first, there may be insufficient available samples to create a viable density estimator; second, sampling of the posterior in (\ref{eq:posterior}) becomes infeasible if the MC must traverse a high-dimensional space.
+
+A concrete example, which motivated this development,
+is when the observable data \vecc\ are the binned 2-point correlation functions of
+cosmic fields derived from a catalog of galaxies; the parameters of
+interest are cosmological quantities such as $\Omega_m, \sigma_8,$
+etc.; and the nuisance parameters $\vecn$
+include the coefficients of some linear expansion of the redshift distributions $n(z)$ of the galaxies being observed:
+\begin{equation}
+  n(z) = \sum_{k=1}^{M} n_k b_k(z).
+  \label{eq:nzbasis}
+\end{equation}
+The $b_k$ are a set of predetermined basis functions for the redshift
+distribution.  In our case there are 10 distinct galaxy populations to be characterized, leading to $O(1000)$ parameters $n_k$ to be considered.
 
@@
 
-For the DES Y3 analyses, \citet{hyperrank} devised a scheme whereby
+Facing this problem for the cosmological analyses of the 3-year data (Y3)
+of the \textit{Dark Energy Survey} (DES),
+\citet{hyperrank} devised a scheme whereby
 the samples of \vecn\ are placed in a grid within some $M$-dimensional
@@
 We describe here an approach that is algebraically similar to that of \citet{hans}, but does not require 2nd condition of Gaussianity for the nuisance prior.
-Our approach is to seek a linear compression of \vecn\ into a lower-dimensional set of parameters \vecu\ that project away variations in \vecn\ that do not influence the likelihood $\likeli.$  Standard density estimators can then be applied to the \vecu\ values implied by the known \vecn\ samples to yield a prior $p(\vecu)$ that can be used for the MC chain of the cosmological posterior.
+Our approach is to seek a linear compression of \vecn\ into a lower-dimensional set of parameters \vecu\ that projects away variations in \vecn\ that do not influence the likelihood $\likeli.$  Standard density estimators can then be applied to the \vecu\ values implied by the known \vecn\ samples to yield a prior $p(\vecu)$ that can be used for the MC chain of the cosmological posterior.
 The model \hatc, and hence $\likeli,$ will be continuous over this low-dimensional \vecu\ space, and marginalization over \vecu\ will yield posterior probabilities very close to marginalization over the original \vecn.
@@
 section we will assume that the \vecn\ vectors have been shifted by
-the mean $\vecn_0 \equiv \left\langle \vecn_\alpha \right\rangle$ so
-that their mean is zero.
+the mean of the samples $\vecn_0 \equiv \left\langle \vecn_\alpha \right\rangle$ so
+that the new \vecn\ has zero mean.
 We seek some function $\hat{\vecn}(\vecu)$ of a lower-dimensional
@@
 $\hat\vecn$ alters the cosmological inference by much less than the
-other uncertainties.  We will implement this by minimizing the
+other uncertainties in the model or data.  We will implement this by minimizing the
 distance in the space $\vecc$ between the
@@
 observations' covariance matrix $\covm_c$ as a metric for the
-distance.  This is equivalent to the $\chi^2$ of the difference
+distance.\footnote{More precisely, the inverse of the covariance
+  matrix is the metric.}  This is equivalent to the $\chi^2$ of the difference
 between the original and compressed models for the data:
@@
                                             \covm_c^{-1}
-                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}\alpha) \right].
+                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}_\alpha) \right].
 \label{eq:chisq}
@@
 \end{equation}
-This quantity is also the Fisher matrix over $\vecn$ giving the information
-provided by the observations $\vecc.$  In many cases this matrix will be
+This quantity is also the Fisher matrix giving the information
+provided by the observations $\vecc$ about the nuisance parameters
+\vecn.  In many cases this matrix will be
 rank-deficient and/or poorly conditioned, since the observables are not
 likely to be very informative on $\vecn$---if they were, we might not be
-concerned with establishing a prior on $\vecn$ to begin with.  We will
-not, however, require the inverse of $\matF$ to proceed.
+concerned with establishing a prior on $\vecn$ to begin with.
+Fortunately, we will not require the inverse of $\matF$ in our algorithm.
 
-The optimization implied by \eqq{\eq:linearized} is the same as in
+The optimization implied by \eqq{eq:linearized} is the same as in
 familiar Principal Components Analysis (PCA), aside from the presence
@@
        & = \trace \left[ \covm_n V_X \proj_{-M} V_X^T \matF V_X
-         \proj_{-M} V_X^T \right].
+         \proj_{-M} V_X^T \right], \\
+  C_n & \equiv \left\langle \vecn \vecn^T \right\rangle.
 \end{align}
@@
 as a lens population and clustering tracer in the Year 6 analysis of
-the \textit{Dark Energy Survey (DES)} galaxy catalogs.  In the baseline case,
+the \textit{Dark Energy Survey (DES)} galaxy catalogs.  Each
+Maglim bin's members are selected with cuts on galaxy fluxes and
+colors in an attempt  to generate a sample that is confined to a
+particular redshift range, as described in \citet{y6gg}.  Once each bin's
+sample is chosen, a combination of 
+photometric techniques \citep{y6pz} and clustering information
+\citep{y6wz} are used to generate samples from the true $n(z)$
+distributions of its members.
+In the baseline case,
 there are 6 cosmological parameters of interest, $\vecq = \{\Omega_m,
@@
 with the tidal field of the mass; and multiplicative errors in the
-measurement of galaxy shear.  The bulk of the nuisance parameters,
-however, are the descriptions of the redshift distribution $n(z)$ of
-galaxies belonging to each of 10 distinct color-based selections
-applied to the galaxy catalog. Each galaxy sample's $n(z)$ distribution is
+measurement of galaxy shear.  The $n(z)$ for each galaxy bin
+is
 described by \eqq{eq:nzbasis} with 80 coefficients spanning $0<z<4$ at
-intervals of $\Delta z=0.05,$  giving a total of 800 nuisance
-parameters.  Each galaxy sample's $n(z)$ is also constrained to
-integrate to unity, which yields 10 linear constraints on these
-nuisance parameters.
+intervals of $\Delta z=0.05.$ If all of these coefficients, for each
+galaxy sample, were allowed to vary, the parameter space for inference would be grow more
+than 10-fold.  The inference would become infeasible even if a
+reliable density estimator over the 80-dimensional space could be
+created from the $\approx3000$ samples available to characterize each
+$n(z).$
 
-Various forms of information from spectroscopic and photometric galaxy
-surveys are used to constrain the $n(z)$ coefficients. As detailed in
-\citet{y6pz}, these generate a set of samples of possible $n(z)$
-coefficients for each galaxy sample.  We show here the results of the
-mode compression process as applied to one of the nine galaxy samples,
-the ??th lowest redshift of 6 lens bins.  We wish to have the
-cosmological Markov chain sample from this
-galaxy sample's 80-vector vector $\vecn$ of coefficients for its
-redshift distribution, being guided by the available 3000??
-equal-probability samples.
+Thus we turn to the modal projection technique herein to reduce the
+dimensionality to those directions in $\vecn$ space in which the samples span a large
+enough range to alter the $\hatc(\vecn)$ at detectable levels.
 
@@
 this sample's positions and the weak gravitational lensing shear
-observed on the samples $1\le s \le 4$ of source galaxies.  A
-covariance matrix spanning all of the experiment's observables is
-available, and for this demonstration we will restrict it to the 30??
-total elements of $\vecc$ that are affected by the $n(z)$ model for
-this bin.
+observed on the samples $1\le s \le 4$ of source galaxies.  The
+derivatives of the model for each observable, and the expected
+covariance, are calculated from theory with the tools described in
+\citet{y6model}.
 
-Figure showing violin plot, before/after.
+\begin{figure}
+  \center
+  \includegraphics[width=0.6\textwidth]{violin.pdf}
+\caption{Violin plots for the redshift probability distribution $n(z)$
+  of galaxies in lens bin 1 for the DES Y6 analysis.  The blue regions
+  show the distributions for the samples of $\vecn$ derived from
+  photometric and clustering information.  The orange violins are for
+  $\vecn$ values drawn defined by (1) compressing the original $\vecn$
+  into 5 modes with coefficients $\vecu,$ then (2) drawing values of
+  $\vecu$ from unit normal distributions and transforming them to
+  match the $\vecu$ 1d distribution of the input samples.  The dashed
+  line connects the mean values of $n(z),$ which are the same for
+  generated samples as for the input samples, by construction.
+  Although the $n(z)$ functions are calculated out to $z=4,$ we
+  truncate this and other plots at lower $z$ to emphasize the
+  lower-redshift regime where this bin's galaxies are primarily found.}
+  \label{fig:violins}
+\end{figure}
 
-Figure showing modes
+We present results for mode-compression sampling of the $n(z)$
+parameters for the lowest-redshift bin.  Figure~\ref{fig:violins}
+plots the distributions of the individual elements of $n(z)$ in the
+input 3000 samples.  As per the procedure desribed in this paper, the
+mean \vecn\ is subtracted from each sample and the covariance
+$\covm_c$ computed.  This is then combined with the
+derivatives and covariance matrices of the observables $\vecc$ to
+calculate the decomposition in \eqq{eq:defG}.
 
-Figure showing $\vecu$ distributions.
+\begin{figure}
+  \center
+  \includegraphics[width=0.6\textwidth]{chiresids.pdf}
+\caption{The size of the $\chi^2$ of modeling error attributable to compressing the
+  $\vecn$ samples down to $M$ modes is plotted vs $M$.  The values
+  drop exponentially with $M,$ and our chosen criterion of
+  $\chi^2<0.15$ is attained with $M=5$ for this bin's $n(z).$}
+  \label{fig:chiresid}
+\end{figure}
 
-Describe 1d normalization mapping.
+Figure~\ref{fig:chiresid} plots the values of unmodelled
+$\chi^2$ vs the dimension $M$ of the compressed space, as per
+\eqq{eq:chiresid}.  This modelling error induced by compression drops
+exponentially with the number of retained modes, and our chosen criterion of
+$\chi^2<0.15$ is attained with $M=5$ modes.
 
-Figure showing histogram of chisq shifts.
+\begin{figure}
+  \center
+  \includegraphics[width=0.6\textwidth]{modes.pdf}
+  \includegraphics[width=0.38\textwidth]{uhist.pdf}
+\caption{At left are plotted the modes, or rows of the decompression
+  matrix $\matD$ in \eqq{UU}, that are derived by our method for $\vecn.$
+ At right are shown the distributions of the coefficients
+  $\vecu$ of the 5 modes that are found by compressing the input
+  samples.  Some of the five are substantially non-Gaussian.}
+\label{fig:modes}
+\end{figure}
+
+The left-hand panel of Figure~\ref{fig:modes} plots the 5 retained
+modes of variation of $n(z),$ \ie\ the rows of the decompression
+matrix $\matD$ in \eqq{UU}.   Recall that each mode will be multiplied by
+a random deviate $u$ with unit variance and added to the mean $\vecn$
+to yield a sample from the prior.
+ Note that although the successively higher-order modes 
+are less detectable by the data, they are not necessarily becoming
+numerically smaller perturbations to $n(z)$.
+
+The right-hand panel of Figure~\ref{fig:modes} plots the distributions
+of the $M=5$ components of the compressed $\vecu$ representations of the
+3000 input samples.  Several of them are substantially non-Gaussian,
+so a unit-normal prior on $\vecu$ will not faithfully represent the
+input samples---a better density estimator is required.  We find in
+this case (and in all other DES cases) that it is sufficient to
+normalize the marginal distributions of the individual $u_i$
+components.  This is done by tabulating a normalizing function $f_i$
+for each mode defined by
+\begin{equation}
+  {\rm CDF}_n\left[ f_i(u_i) \right] = {\rm CDF}(u_i),
+\end{equation}
+where the left side is the cumulative distribution function of the
+unit normal, and the right-hand side is the CDF of the $u_i$ values
+obtained from the input samples.  The functions $f_i$ are bijective
+and can be stored as a splined lookup table.
+Now, the cosmological Markov Chain
+is told that there are five parameters $\widetilde{\vecu}$ that have a
+unit-normal prior $p(\widetilde{\vecu}).$  This vector defines the
+redshift distribution via a 2-step process:
+\begin{align}
+  u_i & =f_i^{-1}(\widetilde{u}_i); \\
+  \vecn & = \matD \vecu.
+\end{align}
+
+\textit{Corner plot of $\widetilde{\vecu})$? ?}
+
+The orange violins in Figure~\ref{fig:violins} show the distributions
+of the $n(z)$ values that are generated by drawing $\widetilde{\vecu}$
+values from a unit Gaussian.  It is clear now that projecting away the
+unobservable fluctuations in $n(z)$ has substantially reduced the
+variance of the function.  To check whether we have achieved our goal
+of leaving the \emph{observable} consequences of the $n(z)$ variation
+unchanged, we calculate the distribution of
+\begin{equation}
+  \chi^2 =  \left[\hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0) \right]^T
+                                            \left[\covm_c^{-1} \hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0)\right]
+\label{eq:chihist}
+\end{equation}
+for the cases (1) the $\vecn$ are drawn from the input samples, and
+(2) are generated using the procedure defined above.  This $\chi^2$
+measures the deviation of the model from that implied by the mean
+vector $\vecn_0$ at some chosen nominal value of cosmological
+parameters $\vecq$ (we also fix the other nuisance parameters of the
+DES model for this test).
+
+Figure~\ref{fig:chihist} shows the results: the $\chi^2$ distribution
+for the input samples is indistinguishable from that of the samples
+generated by our compressed, normalized representation.
+The Figure also shows the $\chi^2$ distribution resulting from samples
+generated by the method used in the DES Y3 analysis \citep{y3pz}.  In
+that case, the $n(z)$ for each galaxy sample was given an \textit{ad
+  hoc} variation of the form
+\begin{equation}
+  n(z) = n_0\left(\frac{z-\Delta\,z}{s}\right),
+\end{equation}
+with $\Delta z$ and $s$ being ``shift'' and ``stretch'' parameters.
+Separable Gaussian priors were assigned to $\Delta z$ and $s$ with
+means of 0 and 1, respectively, and standard deviations that equaled
+the RMS variation in the mean and width of the input $n(z)$ samples.
+In the Figure we can see that the shift-stretch model does not in fact
+reproduce the size of the deviations in the $\vecc$ observables that
+is implied by the original samples.
+
+\begin{figure}
+  \center
+  \includegraphics[width=0.6\textwidth]{chihist.pdf}
+\caption{The histograms show the deviations of the predicted observable $w(\theta)$ quantities in
+  DES Y6 cosmological analysis, as measured by the $\chi^2$ in
+  \eqq{eq:chihist}, as we allow the $n(z)$ parameters to vary.  The
+  ??? histogram shows the variation using the original 3000 samples of
+  $n(z)$ produced by the photometric and clustering redshift studies.
+  The ??? histogram results from
+  drawing 5-dimensional $\widetilde{\vecu}$ values from a unit
+  Gaussian and decompressing them into $n(z)$ realizations using our
+  method.  This lower-dimensional model for the prior on $n(z)$
+  reproduces the original samples' result very well.  By comparison,
+  a \textit{ad hoc} prior for $n(z)$ which reproduces the variations in only the mean
+  and width of the input $n(z)$ samples is less accurate in preserving
+  the fluctuations in the observable quantity under the
+  prior. \textbf{Prelim version of figure.} }
+\label{fig:chihist}
+\end{figure}
 
@@
 \bibitem[Hadzhiyska et al.(2020)]{hans} Hadzhiyska, B., Alonso, D., Nicola, A., et al.\ 2020, \jcap, 2020, 056. doi:10.1088/1475-7516/2020/10/056
+
+\bibitem[Yin et al.(2025)]{y6pz} Yin, B. et al.\ 2025 (PZ paper)
+\bibitem[d'Assignies et al.(2025)]{y6wz} d'Assignies, W. et al.\ 2025 (WZ paper)
+\bibitem[Giannini et al.(2025)]{y6gg} Giannini, G. et al.\ 2025 (lenses paper)
+
 \end{thebibliography}
 
-\bibiten[TBD]{y6pz} Boyan, William, ??? et al.\ 2025 (in preparation)
-Y6 redshift paper(s).
-
 \end{document}

### 2025-02-13 main.tex
some proofing

@@
 vector \vecn\ of nuisance parameters.  
-We wish to characterized the Bayesian posterior probability
+We wish to characterize the Bayesian posterior probability
 \begin{equation}
@@
 The $b_k$ are a set of predetermined basis functions for the redshift
-distribution.  In our case there are 10 distinct galaxy populations to be characterized, leading to $O(1000)$ parameters $n_k$ to be considered.
+distribution.  In our case there are 10 distinct galaxy populations to
+be characterized, each with $n(z)$ tabulated at $\approx100$ values, leading to $O(1000)$ parameters $n_k$ to be considered.
 
@@
   function of \vecu, are discontinuous.
-Various strategies are proposed to assign the $\vecn_\alpha$ to the grid points in
+Various strategies are proposed by \citet{hyperrank} to assign the $\vecn_\alpha$ to the grid points in
 $\mathcal{H}$ based on summary statistics, to reduce the
-discontinuities.  But the function is never smooth.
+discontinuities---but the function is never smooth.
 As a consequence, many MC samplers become quite inefficient in
@@
 
-We describe here an approach that is algebraically similar to that of \citet{hans}, but does not require 2nd condition of Gaussianity for the nuisance prior.
+We describe here an approach that is algebraically similar to that of
+\citet{hans}, but does not require 2nd condition of Gaussianity for
+the nuisance prior, and is likely to be a good approximation even when
+the first and third conditions are not grossly violated.
 Our approach is to seek a linear compression of \vecn\ into a lower-dimensional set of parameters \vecu\ that projects away variations in \vecn\ that do not influence the likelihood $\likeli.$  Standard density estimators can then be applied to the \vecu\ values implied by the known \vecn\ samples to yield a prior $p(\vecu)$ that can be used for the MC chain of the cosmological posterior.
@@
 in order of decreasing eigenvalue $\lambda^G_i.$ The
-elements surviving the project functions yield
+elements surviving the projection $P_{-M}$ yield
 \begin{equation}
@@
   \chi^2 =  \left[\hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0) \right]^T
-                                            \left[\covm_c^{-1} \hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0)\right]
+                                           \covm_c^{-1}  \left[ \hatc(\vecq,\vecn) - \hatc(\vecq,\vecn_0)\right]
 \label{eq:chihist}
@@
 the RMS variation in the mean and width of the input $n(z)$ samples.
+The shift-stretch model essentially compresses the samples of $n(z)$
+into two parameters, their mean and standard deviation, and executes
+an \textit{ad hoc} reconstruction based on those parameters.
 In the Figure we can see that the shift-stretch model does not in fact

### 2025-02-17 main.tex
Full text and figures, IR candidate

@@
 
-\author{Troxel,William,Boyan,Giulia,Alex Amon,Alex Alarcon,\ldots}
+\author{William d'Assignies}
+\author{Michael Troxel}
+\author{Alex Alarcon}
+\author{Alex Amon}
+\author{Giulia Giannini}
+\author{Boyan Yin}
 
@@
 
-The scenario we address here is when \emph{the prior $p(\vecn)$ is itself known only from a set of samples of $\vecn$ from this distribution.} Most MC samplers require that the posterior be an evaluable function of any value of the parameters, and it is the general task of density estimators to convert the samples of $\vecn$ into an evaluable $p(\vecn).$  But when \vecn\ is of high dimension, two problems arise: first, there may be insufficient available samples to create a viable density estimator; second, sampling of the posterior in (\ref{eq:posterior}) becomes infeasible if the MC must traverse a high-dimensional space.
+The scenario we address here is when \emph{the prior $p(\vecn)$ is
+  itself known only from a set of samples of $\vecn$ from this
+  distribution.} Most MC samplers require that the posterior be an
+evaluable function of any value of the parameters---and hence of the
+priors.
+It is the general task of density estimators to convert the samples of $\vecn$ into an evaluable $p(\vecn).$  But when \vecn\ is of high dimension, two problems arise: first, there may be insufficient available samples to create a viable density estimator; second, sampling of the posterior in (\ref{eq:posterior}) becomes infeasible if the MC must traverse a high-dimensional space.
 
-A concrete example, which motivated this development,
+A concrete example, which motivated this paper's work,
 is when the observable data \vecc\ are the binned 2-point correlation functions of
 cosmic fields derived from a catalog of galaxies; the parameters of
-interest are cosmological quantities such as $\Omega_m, \sigma_8,$
+interest are cosmological quantities such as the dark matter density
+$\Omega_m,$ the amplitude of density fluctuations $\sigma_8,$
 etc.; and the nuisance parameters $\vecn$
@@
 \begin{equation}
-  n(z) = \sum_{k=1}^{M} n_k b_k(z).
+  n(z) = \sum_{k=1}^{N} n_k b_k(z).
   \label{eq:nzbasis}
@@
 The $b_k$ are a set of predetermined basis functions for the redshift
-distribution.  In our case there are 10 distinct galaxy populations to
-be characterized, each with $n(z)$ tabulated at $\approx100$ values, leading to $O(1000)$ parameters $n_k$ to be considered.
+distribution.  In our case of analyzing the data from the  \textit{Dark Energy Survey} (DES), there are 10 distinct galaxy populations to
+be characterized, each with $n(z)$ tabulated at $\approx100$ values, leading to $N=O(1000)$ parameters $n_k$ to be considered.
 
@@
 
-Facing this problem for the cosmological analyses of the 3-year data (Y3)
-of the \textit{Dark Energy Survey} (DES),
+Facing this problem for the cosmological analyses of the 3-year data
+(Y3) from DES,
 \citet{hyperrank} devised a scheme whereby
@@
 As a consequence, many MC samplers become quite inefficient in
-sampling of the cosmological posterior.
-In particular, samplers such as \textsc{MultiNest} that assume continuity are rendered nearly non-functional.
-As a result, the Y3 cosmological priors could not be evaluated with this method.
-Instead, the \vecn\ samples were not used, and an \textit{ad hoc} $p(\vecn)$ was adopted which allowed only shifts and dilations of the mean $n(z)$ of the \vecn\ samples.
+sampling of the cosmological posterior.  In particular, samplers such
+as \textsc{MultiNest} that assume continuity are rendered nearly
+non-functional.  As a result, the Y3 cosmological priors could not be
+evaluated with this method.  Instead, the \vecn\ samples were not
+used, and an \textit{ad hoc} $p(\vecn)$ was adopted which allowed only
+shifts and dilations of the mean $n(z)$ of the \vecn\ samples [see \eqq{eq:zs}].
 
@@
 \end{align}
-This optimization is easier if we transform first the systematic variables to
+This optimization is easier if we first transform the systematic variables to
 $\vecn^\prime =\matT\vecn$ such that $\covm_{n^\prime}=\matI$, \ie\
@@
 in all of the input samples $\vecn_\alpha,$ so that the $\vecn$ are
-confined to a subspace---the reconstructed $\hat\vecn$ should also be.
+confined to a subspace---therefore the reconstructed $\hat\vecn$ should also be.
 The compressed
@@
 redshift distribution of one of the bins of ``Maglim''  galaxies used
-as a lens population and clustering tracer in the Year 6 analysis of
-the \textit{Dark Energy Survey (DES)} galaxy catalogs.  Each
+as a lens population and clustering tracer in the Year 6 (Y6) analysis of
+the DES galaxy catalogs.  Each
 Maglim bin's members are selected with cuts on galaxy fluxes and
@@
 photometric techniques \citep{y6pz} and clustering information
-\citep{y6wz} are used to generate samples from the true $n(z)$
-distributions of its members.
+\citep{y6wz} is used to generate samples from the posterior
+distribtuion of  the $n(z)$ distribution of the bin members, conditioned
+on the photometric and clustering data.
+
 In the baseline case,
@@
 
-The observable quantities $\vecc$ whose modeling depends upon the these nuisance
-parameters are the angular autocorrelation $w(\theta)$ of the sample
-members, and the cross-correlations $\gamma_t^{(s)}(\theta)$ between
-this sample's positions and the weak gravitational lensing shear
-observed on the samples $1\le s \le 4$ of source galaxies.  The
+The observable quantities $\vecc$ whose modeling depends upon these nuisance
+parameters are the angular autocorrelation $w(\theta)$ of the 
+galaxies, and the cross-correlations $\gamma_t^{(s)}(\theta)$ between
+these galaxies' positions and the weak gravitational lensing shear
+observed from groups $1\le s \le 4$ of source galaxies.  The
 derivatives of the model for each observable, and the expected
 covariance, are calculated from theory with the tools described in
-\citet{y6model}.
+\citet{y6model} and the \texttt{Cosmosis} software\footnote{\url{https://cosmosis.readthedocs.io/en/latest/}} \citet{cosmosis}.
 
@@
   \center
-  \includegraphics[width=0.6\textwidth]{violin.pdf}
+  \includegraphics[width=0.6\textwidth]{bin4_violins.pdf}
 \caption{Violin plots for the redshift probability distribution $n(z)$
-  of galaxies in lens bin 1 for the DES Y6 analysis.  The blue regions
+  of galaxies in lens bin 4 for the DES Y6 analysis.  The blue regions
   show the distributions for the samples of $\vecn$ derived from
@@
   $\vecn$ values drawn defined by (1) compressing the original $\vecn$
-  into 5 modes with coefficients $\vecu,$ then (2) drawing values of
-  $\vecu$ from unit normal distributions and transforming them to
-  match the $\vecu$ 1d distribution of the input samples.  The dashed
+  into 3 modes with coefficients $\vecu,$ then (2) drawing values of
+  $\widetilde u_i$ from unit normal distributions and transforming each
+  component $u_i$ to match the 1d distribution of the input samples' $u_i.$.  The dashed
   line connects the mean values of $n(z),$ which are the same for
@@
 We present results for mode-compression sampling of the $n(z)$
-parameters for the lowest-redshift bin.  Figure~\ref{fig:violins}
+parameters for redshift bin 4 of the DES Y6 lens galaxies.  Figure~\ref{fig:violins}
 plots the distributions of the individual elements of $n(z)$ in the
@@
   \center
-  \includegraphics[width=0.6\textwidth]{chiresids.pdf}
-\caption{The size of the $\chi^2$ of modeling error attributable to compressing the
+  \includegraphics[width=0.49\textwidth]{chiresids.pdf}
+   \includegraphics[width=0.49\textwidth]{modes.pdf}
+\caption{At left: The size of the $\chi^2$ of modeling error attributable to compressing the
   $\vecn$ samples down to $M$ modes is plotted vs $M$.  The values
   drop exponentially with $M,$ and our chosen criterion of
-  $\chi^2<0.15$ is attained with $M=5$ for this bin's $n(z).$}
-  \label{fig:chiresid}
-\end{figure}
+  $\chi^2<0.025$ is attained with $M=3$ for this bin's
+  $n(z).$
+  At right: The modes of variation $U_i(z),$, \ie\ the rows of the decompression
+  matrix $\matD$ in \eqq{eq:DE} are plotted vs redshift.  Each mode is
+  multiplied by a unit-variance stochastic coefficient $u_i$, then
+  they are summed with the mean $\bar\vecn(z),$ to form an $n(z)$
+  sample.  Higher-numbered modes have
+  observable consequences of decreasing statistical significance.}
+  \label{fig:chiresid}\end{figure}
 
-Figure~\ref{fig:chiresid} plots the values of unmodelled
+The left plot in Figure~\ref{fig:chiresid} shows the values of unmodelled
 $\chi^2$ vs the dimension $M$ of the compressed space, as per
@@
 exponentially with the number of retained modes, and our chosen criterion of
-$\chi^2<0.15$ is attained with $M=5$ modes.
+$\chi^2<0.025$ is attained with $M=3$ modes for this bin.  The modes
+$U_i(z),$ \ie\ the rows of the decompression matrix $\matD,$ are
+plotted on the right of Figure~\ref{fig:chiresid}.  Recall that each mode will be multiplied by
+a random deviate $u$ with unit variance and added to the mean $\vecn$
+to yield a sample from the prior.
 
@@
   \center
-  \includegraphics[width=0.6\textwidth]{modes.pdf}
-  \includegraphics[width=0.38\textwidth]{uhist.pdf}
-\caption{At left are plotted the modes, or rows of the decompression
-  matrix $\matD$ in \eqq{UU}, that are derived by our method for $\vecn.$
- At right are shown the distributions of the coefficients
-  $\vecu$ of the 5 modes that are found by compressing the input
-  samples.  Some of the five are substantially non-Gaussian.}
-\label{fig:modes}
+  \includegraphics[width=0.7\textwidth]{bin4_corner.pdf}
+  \caption{In orange is a corner plot of the distribution of the mode coefficients,
+    \ie\ the elements of $\vecu = \matE \vecn.$
+  of the input samples after encoding.  The coefficients, especially 
+  $u_1,$ are significantly skewed so a normal distribution would be an
+  inaccurate model.  Instead we model each $u_i$ as an
+  ``denormalizing'' function of a unit-normal variable, as per
+  \eqq{eq:denorm}.  The blue histograms and contours show the
+  distributions obtained using this element-by-element transformation
+  technique, which is seen to accurately reproduce the distribution of
+  the input samples.}
+\label{fig:corner}
 \end{figure}
 
-The left-hand panel of Figure~\ref{fig:modes} plots the 5 retained
-modes of variation of $n(z),$ \ie\ the rows of the decompression
-matrix $\matD$ in \eqq{UU}.   Recall that each mode will be multiplied by
-a random deviate $u$ with unit variance and added to the mean $\vecn$
-to yield a sample from the prior.
- Note that although the successively higher-order modes 
-are less detectable by the data, they are not necessarily becoming
-numerically smaller perturbations to $n(z)$.
-
-The right-hand panel of Figure~\ref{fig:modes} plots the distributions
-of the $M=5$ components of the compressed $\vecu$ representations of the
-3000 input samples.  Several of them are substantially non-Gaussian,
+Figure~\ref{fig:corner} plots the distributions
+of the $M=3$ components of the compressed $\vecu$ representations of the
+3000 input samples.  Some are substantially non-Gaussian,
 so a unit-normal prior on $\vecu$ will not faithfully represent the
@@
   {\rm CDF}_n\left[ f_i(u_i) \right] = {\rm CDF}(u_i),
+\label{eq:denorm}
 \end{equation}
@@
 \end{align}
-
-\textit{Corner plot of $\widetilde{\vecu})$? ?}
 
@@
 
+\begin{figure}
+  \center
+  \includegraphics[width=0.6\textwidth]{bin4_chisq.pdf}
+\caption{The histograms show the deviations of the predicted observable $w(\theta)$ quantities in
+  DES Y6 cosmological analysis, as measured by the $\chi^2$ in
+  \eqq{eq:chihist}, as we allow the $n(z)$ parameters to vary.  The
+  shaded green histogram shows the variation using the original 3000 samples of
+  $n(z)$ produced by the photometric and clustering redshift studies.
+  The dashed yellow histogram results from
+  drawing 3-dimensional $\widetilde{\vecu}$ values from a unit
+  normal and decompressing them into $n(z)$ realizations using our
+  method.  This lower-dimensional model for the prior on $n(z)$
+  reproduces the original samples' result very well.  By comparison,
+  the dash-dotted red histogram generates samples of $n(z)$ using
+  the \textit{ad hoc} method of \eqq{eq:zs}, which reproduces the
+  observable behavior of the original samples very well.  This method's two
+  parameters $z$ and $s$ are given priors to match the distributions
+  of the mean of $n(z)$ over $z$, and the standard deviation of $z$,
+  present in the input samples.  The \textit{ad hoc} method produces
+  $n(z)$ fluctuations with $\approx20\%$ larger $\chi^2$ from the mean,
+  on average, than the input distribution has.}
+\label{fig:chihist}
+\end{figure}
+
 Figure~\ref{fig:chihist} shows the results: the $\chi^2$ distribution
@@
   n(z) = n_0\left(\frac{z-\Delta\,z}{s}\right),
+  \label{eq:zs}
 \end{equation}
@@
 reproduce the size of the deviations in the $\vecc$ observables that
-is implied by the original samples.
+is implied by the original samples; the mean $\chi^2$ induced by the
+shift-stretch samples is $\approx20\%$ larger than the input samples.  
+This mismatch indicates that some cosmologically relevant information in the
+original samples is being lost, or that the shift-stretch samples are
+imposing cosmological constraints that are not present in the original
+samples.
 
-\begin{figure}
-  \center
-  \includegraphics[width=0.6\textwidth]{chihist.pdf}
-\caption{The histograms show the deviations of the predicted observable $w(\theta)$ quantities in
-  DES Y6 cosmological analysis, as measured by the $\chi^2$ in
-  \eqq{eq:chihist}, as we allow the $n(z)$ parameters to vary.  The
-  ??? histogram shows the variation using the original 3000 samples of
-  $n(z)$ produced by the photometric and clustering redshift studies.
-  The ??? histogram results from
-  drawing 5-dimensional $\widetilde{\vecu}$ values from a unit
-  Gaussian and decompressing them into $n(z)$ realizations using our
-  method.  This lower-dimensional model for the prior on $n(z)$
-  reproduces the original samples' result very well.  By comparison,
-  a \textit{ad hoc} prior for $n(z)$ which reproduces the variations in only the mean
-  and width of the input $n(z)$ samples is less accurate in preserving
-  the fluctuations in the observable quantity under the
-  prior. \textbf{Prelim version of figure.} }
-\label{fig:chihist}
-\end{figure}
+\section{Summary}
+  We present a linear dimensionality reduction technique that has the
+  aim of making it feasible to produce continuous density estimators
+  for nuisance parameters that are known only from a set of samples in
+  a high-dimensional space $\vecn.$  This is essentially a principle
+  components analysis that is adjusted to separate contributions to
+  the detectable consequences of $\vecn$ rather than contributions to
+  the Euclidean norm of $\vecn.$   This method enables a rigorous
+  marginalization over the distribution of $\vecn$ even in for
+  non-Gaussian distributions, as long as the derivatives of the
+  log-likelihood of the data with respect to $\vecn$ are nearly
+  constant over the posterior domain.  The mode-projection technique
+  is also useful as a method of projecting away irrelevant
+  noise fluctuations in $\vecn.$
+
+  We have shown the utility of this technique for the case when
+  $\vecn$ represents the redshift distributions of galaxies in DES.
+  The resulting modeled distribution of $n(z)$'s is more faithful to
+  the original sample than the previous \textit{ad hoc} method of
+  shifting and stretching the mean $n(z)$ to mimic the samples'
+  range.  While this difference has minimal consequence for the
+  posterior cosmological parameter estimates in the DES Y6 analysis,
+  the mode-projection method is better motivated and will produce more
+  accurate posteriors in future experiments where the $n(z)$
+  prior's uncertainty is a dominant contributor to the cosmological posterior.
 
   \begin{acknowledgments}
-
 G.M.B. acknowledges support from NSF grant AST-2205808 and \ldots.
@@
 %\bibliographystyle{aasjournal}
+\bibitem[d'Assignies et al.(2025)]{y6wz} d'Assignies, W. et al.\ 2025 (WZ paper)
+
 \bibitem[Cordero et al.(2022)]{hyperrank} Cordero, J.~P., Harrison, I., Rollins, R.~P., et al.\ 2022, \mnras, 511, 2170. doi:10.1093/mnras/stac147
+
+\bibitem[Giannini et al.(2025)]{y6gg} Giannini, G. et al.\ 2025 (lenses paper)
 
@@
 
+\bibitem[Myles et al.(2021)]{y3pz}
+  Myles, J. et al.\ 2021, \mnras, 505, 4249,
+  doi:10.1093/mnras/stab1515
+
 \bibitem[Yin et al.(2025)]{y6pz} Yin, B. et al.\ 2025 (PZ paper)
-\bibitem[d'Assignies et al.(2025)]{y6wz} d'Assignies, W. et al.\ 2025 (WZ paper)
-\bibitem[Giannini et al.(2025)]{y6gg} Giannini, G. et al.\ 2025 (lenses paper)
+
+\bibitem[Zuntz et al.(2015)]{cosmosis}
+  Zuntz, J. et al. 2015, Astronomy and Computing, 12, 45,
+  doi:10.1016/j.ascom.2015.05.005
+
+\bibitem[??? et al.(2025)]{y6model}
+  Y6 modeling paper ???
 

### 2025-03-22 main.tex
CWR version

@@
 
+\newcommand{\ed}[1]{{\color{red}{#1}}}
+\newcommand\gary[1]{{\color{red} \{\textbf{GMB}: #1\}}}
+
 \newcommand{\ie}{\textit{i.e.}}
@@
 \newcommand{\eqq}[1]{Equation~(\ref{#1})}
-\newcommand\gary[1]{{\color{red} \{\textbf{GMB}: #1\}}}
 %\tabletypesize{\footnotesize}
@@
 \newcommand{\vecu}{\ensuremath{\mathbf{u}}}
+\newcommand{\vecU}{\ensuremath{\mathbf{U}}}
 \newcommand{\vecv}{\ensuremath{\mathbf{v}}}
@@
 \author{Boyan Yin}
+\author{Builders\ldots} %%%%
 
@@
 parameters.  This posterior is complex enough that it requires
-approximation by the output of a Markov chain (MC) wandering across the space $(\vecq,\vecn).$
+approximation by the output of a Markov Chain (MC) wandering across the space $(\vecq,\vecn).$
 
@@
   distribution.} Most MC samplers require that the posterior be an
-evaluable function of any value of the parameters---and hence of the
-priors.
+evaluable function of any value of the parameters.
 It is the general task of density estimators to convert the samples of $\vecn$ into an evaluable $p(\vecn).$  But when \vecn\ is of high dimension, two problems arise: first, there may be insufficient available samples to create a viable density estimator; second, sampling of the posterior in (\ref{eq:posterior}) becomes infeasible if the MC must traverse a high-dimensional space.
@@
 cosmic fields derived from a catalog of galaxies; the parameters of
-interest are cosmological quantities such as the dark matter density
+interest are cosmological quantities such as the  matter density
 $\Omega_m,$ the amplitude of density fluctuations $\sigma_8,$
@@
 The $b_k$ are a set of predetermined basis functions for the redshift
-distribution.  In our case of analyzing the data from the  \textit{Dark Energy Survey} (DES), there are 10 distinct galaxy populations to
-be characterized, each with $n(z)$ tabulated at $\approx100$ values, leading to $N=O(1000)$ parameters $n_k$ to be considered.
+distribution.
+In our case of analyzing the data from the  \textit{Dark Energy Survey} (DES), there are 10 distinct samples of galaxies---each designed to prefer galaxies in a limited range of redshift---which we can index by $s.$ Each has its own $n_s(z)$ to be characterized by coefficients $n_{sk}$ at $\approx100$ values of $k$, leading to $N=O(1000)$ parameters $n_{sk}$ to be considered.
+The vector $\vecn$ of nuisance parameters would be the concatenation of all the $n_{sk}.$ For clarity, we will still write this as $\vecn=\{n_1,n_2,\ldots,n_N\}$ and demonstrate the method with a single galaxy sample's $n(z)$.
 
@@
 
-A more rigorous and extremely efficient method of marginalizing over high-dimensional nuisance parameters was proposed by \citet{hans}, for the case where the following restrictions apply:
+A more rigorous and extremely efficient method of marginalizing over high-dimensional nuisance parameters was 
+described by \citet{bridle02} and reprised by \citet{hans} for the $n(z)$ application, for the case where the following restrictions apply:
 \begin{enumerate}
 \item The likelihood of the observable \vecc\ is normal, $\vecc \sim \mathcal{N}( \hatc, \covm_c),$ with $\covm_c$ fixed.
-\item The prior $p(\vecn)$ can also be assumed to be normal, with a mean and covariance matrix $\covm_n$ that in our case could be assigned from the mean and covariance of the samples of \vecn\ we are given.
+\item The prior $p(\vecn)$ can also be assumed to be normal, with a mean taken to be $\bar\vecn = \left\langle\vecn\right\rangle$ and covariance matrix taken to be $\covm_n=\left\langle(\vecn-\bar\vecn)(\vecn-\bar\vecn)^T\right\rangle$ using the samples of \vecn\ we are given.
 \item The model \hatc\ can be linearized in \vecn\ about fiducial values $\vecq_0, \vecn_0$ without loss of accuracy exceeding measurement errors, with the derivatives independent of \vecq.
 \end{enumerate}
-Under these conditions, \citet{hans} show that the marginalization over \vecn\ is equivalent to adding terms to $\covm_c,$ such that any MC process need not sample \vecn\ at all.
+Under these conditions, the marginalization over \vecn\ is equivalent to adding terms to $\covm_c,$ such that any MC process need not sample \vecn\ at all.
 
-We describe here an approach that is algebraically similar to that of
-\citet{hans}, but does not require 2nd condition of Gaussianity for
-the nuisance prior, and is likely to be a good approximation even when
+We describe here an approach that is algebraically similar to this analytic marginalization,
+but does not require the 2nd condition of Gaussianity for
+the nuisance prior, and is likely to be a good approximation when
 the first and third conditions are not grossly violated.
@@
 that the new \vecn\ has zero mean.
-We seek some function $\hat{\vecn}(\vecu)$ of a lower-dimensional
+We seek some function $\hat{\vecn}(\vecu)$ of a much lower-dimensional
 vector \vecu\ which can be substituted for \vecn\ and yield nearly the
@@
 samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha.$
-This means we want a map $\vecn_\alpha\rightarrow
-\vecu_\alpha \rightarrow \hat{\vecn}_\alpha$ such that replacing $\vecn$ with
+This means we want maps $\vecn_\alpha\rightarrow
+\vecu_\alpha \rightarrow \hat{\vecn}_\alpha,$ with ${\rm dim}(\vecn)={\rm dim}(\hat{\vecn})=N$ and ${\rm dim}({\vecu})=M\ll N.$ We wish to find maps such that replacing $\vecn$ with
 $\hat\vecn$ alters the cosmological inference by much less than the
@@
 We next assume that the compression is linear, $\hat\vecn=\matX\vecn,$
-for some matrix that is idempotent ($\matX\matX = \matX$).  If this is
-true, then the contribution to \eqq{eq:chisq} at lowest order in
-$\vecn-\hat\vecn$ is
+for some $N\times N$ matrix that is idempotent ($\matX\matX = \matX$).  We will further linearize the dependence of $\hat\vecc$ on $\vecn,$ specifically assuming that (in scalar notation) 
+\begin{equation}
+    \frac{\partial^2 \hat c}{\partial n^2} n \ll \frac{\partial \hat c}{\partial n}
+\end{equation}
+over the full range of variation of $\vecn.$  
+With these two assumptions, \eqq{eq:chisq} becomes
 \begin{equation}
   \left\langle \chi^2 \right\rangle = \frac{1}{N_\alpha} \sum_\alpha
-  \left[ (\matI-\matX)\vecn_\alpha\right]^T \matF  \left[ (\matI-\matX)\vecn_\alpha\right]
+  \left[ (\matI-\matX)\vecn_\alpha\right]^T \matF  \left[ (\matI-\matX)\vecn_\alpha\right].
   \label{eq:linearized}
 \end{equation}
-where we use the Jacobian matrix of the model $\hatc$ to define
+We use the Jacobian matrix of the model $\hatc$ to define
 \begin{equation}
   \matF \equiv
-  \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0, \vecn_0}
+  \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0, \vecn_0}^T
   \covm_c^{-1} \left[\frac{\partial\hatc}{\partial\vecn}\right]_{\vecq_0,
-    \vecn_0}^T.
+    \vecn_0}.
 \label{eq:fisher}
@@
 Since $\matX$ is idempotent, we can write
-\begin{align}
-  \matX & = V_X \proj_M V_X^T, \\
-  \matY \equiv \matI-\matX & = V_X \proj_{-M} V_X^T,
-\end{align}
+\begin{equation}
+  \matX = V_X \proj_M V_X^T,
+\end{equation}
 where $V_X$ is unitary and the projection matrix $\proj_M$ is defined as
@@
 \end{equation}
-and $\proj_{-M}=\matI-\proj_M.$  For a chosen rank $M$ of the
+It is also useful to define
+\begin{equation}
+  \matY \equiv \matI-\matX  = V_X \proj_{-M} V_X^T,
+\end{equation}
+with $\proj_{-M}=\matI-\proj_M.$
+
+For a chosen rank $M$ of the
 transformation matrix $X$, our task becomes to identify the
@@
     \proj_{-M} V_{X^\prime}^T \right] \\
-  & = \trace \left[ \proj_{-M} V_{X^\prime}^T \matV_G \matLam_G \matV_G^T
+  & = \trace \left[ \proj_{-M} V_{X^\prime}^T \matG 
     V_{X^\prime} \proj_{-M} \right],
-  \label{eq:mintr}\\
-  \matG \equiv \left(T^{-1}\right)^T \matF   T^{-1} & = \matLam_n^{1/2} \matV_n^T
+  \label{eq:mintr}
+\end{align}
+where we have defined the transformed Fisher matrix
+\begin{equation}
+  \matG \equiv \left(T^{-1}\right)^T \matF   T^{-1} = \matLam_n^{1/2} \matV_n^T
   \matF \matV_n \matLam_n^{1/2} = \matV_G \matLam_G \matV_G^T.
   \label{eq:defG}
-\end{align}
-The last line defines a transformed Fisher matrix $\matG$ and its
-(non-negative) eigensystem.  Examination of \eqq{eq:mintr} reveals that this quantity
-is minimized if the elements of the unitary matrices satisfy
+\end{equation}
+The right-hand side defines the eigensystem of $G$, with $\Lambda_G={\rm diag}(\lambda^G_1,\ldots,\lambda^G_N),$ and $\lambda^G_i\ge0.$  \eqq{eq:mintr} can now be rewritten as
+\begin{equation}
+\left\langle \chi^2\right\rangle = \sum_{i>M} \left(\matV\lambda_G \matV^T\right)_{ii}
+\end{equation}
+where $\matV=\matV^T_{X^\prime}\matV_G$ is unitary.  This expression must be at least as large as sum of the $N-M$ smallest $\lambda^G_i$, and that minimum is attained if
 $\matV_{X^\prime}^T \matV_G = \matI \; \Rightarrow \; \matV_{X^\prime} = \matV_G,$ and the eigensystem of $G$ is placed
-in order of decreasing eigenvalue $\lambda^G_i.$ The
+in order of decreasing eigenvalues $\lambda^G_i.$ The
 elements surviving the projection $P_{-M}$ yield
@@
 \begin{enumerate}
-  \item Obtain the Fisher matrix $\matF$ of the system defined in
-    \eqq{eq:fisher} using derivatives about the mean sample $\vecn_0,$
+  \item Obtain the mean $\vecn_0$ of the input samples, the Fisher matrix $\matF$ of the system defined in
+    \eqq{eq:fisher} using derivatives about $\vecn_0,$ 
     plus the covariance matrix $\covm_n$ of the samples.
@@
 marginalization.  One can, for example, examine the posterior
-distributions of \vecu\ and see if they are at the edges of the prior,
-which would potentially indicate an inconsistency between the data and
+distributions of \vecu\ to see how the data have constrained the prior.
+For example, if the cosmological posterior for $\vecu$ is at the edge of the prior, this would potentially indicate an inconsistency between the data and
 the prior.
@@
 as a lens population and clustering tracer in the Year 6 (Y6) analysis of
-the DES galaxy catalogs.  Each
-Maglim bin's members are selected with cuts on galaxy fluxes and
+the DES galaxy catalogs.  Each of these Maglim ``lens bins''
+is selected with cuts on galaxy fluxes and
 colors in an attempt to generate a sample that is confined to a
@@
 
-In the baseline case,
+In the simplest case,
 there are 6 cosmological parameters of interest, $\vecq = \{\Omega_m,
 \Omega_b, \sigma_8, h, n_s, m_\nu\}.$   The nuisance vector $\vecn$
-has 5--10 parameters in each of the following categories: galaxy
+has $\approx 5$  parameters in each of the following categories: galaxy
 biases with respect to matter; intrinsic alignments of galaxy shapes
-with the tidal field of the mass; and multiplicative errors in the
-measurement of galaxy shear.  The $n(z)$ for each galaxy bin
+with the tidal field of the mass; magnification coefficients; and multiplicative errors in the
+measurement of galaxy shear, for a total of 24 non-redshift parameters.  The $n(z)$ for each galaxy bin
 is
@@
 intervals of $\Delta z=0.05.$ If all of these coefficients, for each
-galaxy sample, were allowed to vary, the parameter space for inference would be grow more
-than 10-fold.  The inference would become infeasible even if a
+of the 10 galaxy selections, were allowed to vary, the parameter space for inference would grow from 24 to 824 dimensions.
+The inference would become infeasible even if a
 reliable density estimator over the 80-dimensional space could be
-created from the $\approx3000$ samples available to characterize each
+created from the $O(10^4)$ samples available to characterize each
 $n(z).$
 
-Thus we turn to the modal projection technique herein to reduce the
+We hence turn to the modal projection technique herein to reduce the
 dimensionality to those directions in $\vecn$ space in which the samples span a large
@@
 covariance, are calculated from theory with the tools described in
-\citet{y6model} and the \texttt{Cosmosis} software\footnote{\url{https://cosmosis.readthedocs.io/en/latest/}} \citet{cosmosis}.
+\citet{y6model} and the \texttt{Cosmosis} software\footnote{\url{https://cosmosis.readthedocs.io/en/latest/}} \citep{cosmosis}.
 
@@
   photometric and clustering information.  The orange violins are for
-  $\vecn$ values drawn defined by (1) compressing the original $\vecn$
-  into 3 modes with coefficients $\vecu,$ then (2) drawing values of
-  $\widetilde u_i$ from unit normal distributions and transforming each
-  component $u_i$ to match the 1d distribution of the input samples' $u_i.$.  The dashed
+  $\vecn$ values drawn defined by (1) subtracting the mean $\bar\vecn$; (2) compressing these $\vecn$
+  into 3 modes with coefficients $\vecu;$ (2) drawing values of
+  $\widetilde{u}_i$ from unit normal distributions; (3) transforming each
+  component $\widetilde u_i$ to match the 1d distribution of the input samples' $u_i$; (4) decompressing the transformed $u_i$ draws back into full-length $\vecn$ samples; finally (5) restoring the mean $\bar\vecn.$  The dashed
   line connects the mean values of $n(z),$ which are the same for
   generated samples as for the input samples, by construction.
-  Although the $n(z)$ functions are calculated out to $z=4,$ we
+The compression substantially lowers the variance of $n(z)$ at individual values of $z$ without significantly altering the variation of cosmological signals that the entire $n(z)$ predicts.
+  [Although the $n(z)$ functions are calculated out to $z=4,$ we
   truncate this and other plots at lower $z$ to emphasize the
-  lower-redshift regime where this bin's galaxies are primarily found.}
+  lower-redshift regime where this bin's galaxies are primarily found.]}
   \label{fig:violins}
@@
 plots the distributions of the individual elements of $n(z)$ in the
-input 3000 samples.  As per the procedure desribed in this paper, the
+input 3000 samples.  As per the procedure described in this paper, the
 mean \vecn\ is subtracted from each sample and the covariance
-$\covm_c$ computed.  This is then combined with the
+$\covm_n$ computed.  This is then combined with the
 derivatives and covariance matrices of the observables $\vecc$ to
@@
 \caption{At left: The size of the $\chi^2$ of modeling error attributable to compressing the
-  $\vecn$ samples down to $M$ modes is plotted vs $M$.  The values
-  drop exponentially with $M,$ and our chosen criterion of
+  $\vecn$ samples down to $M$ modes is plotted vs $M$.
+The $M=0$ point shows the modelling error from holding $n(z)$ fixed at its mean, and $M\ge1$ values drop exponentially as we use more modes to reconstruct $n(z).$ Our chosen criterion of
   $\chi^2<0.025$ is attained with $M=3$ for this bin's
   $n(z).$
-  At right: The modes of variation $U_i(z),$, \ie\ the rows of the decompression
-  matrix $\matD$ in \eqq{eq:DE} are plotted vs redshift.  Each mode is
+  At right: The modes of variation $U_i(z),$ \ie\ the rows of the decompression
+  matrix $\matD$ in \eqq{eq:DE}, are plotted vs redshift.  Each of
+  modes 0,1,2 is
   multiplied by a unit-variance stochastic coefficient $u_i$, then
@@
   sample.  Higher-numbered modes have
-  observable consequences of decreasing statistical significance.}
+  observable consequences of decreasing statistical significance.
+Mode 7 is plotted as the dashed blue line as an example of what is projected out of $n(z)$; even though its typical amplitude in the input data is larger than modes 1 or 2, its oscillatory behavior does not lead to measurable changes in the cosmological statistics.}
   \label{fig:chiresid}\end{figure}
@@
 \eqq{eq:chiresid}.  This modelling error induced by compression drops
-exponentially with the number of retained modes, and our chosen criterion of
-$\chi^2<0.025$ is attained with $M=3$ modes for this bin.  The modes
-$U_i(z),$ \ie\ the rows of the decompression matrix $\matD,$ are
-plotted on the right of Figure~\ref{fig:chiresid}.  Recall that each mode will be multiplied by
-a random deviate $u$ with unit variance and added to the mean $\vecn$
-to yield a sample from the prior.
+exponentially with the number of retained modes.
+We have somewhat arbitrarily chosen a threshold of $\chi^2<0.025$ for each galaxy sample in order to keep the total impact from modal compression $\ll 1.$ This is attained with $M=3$ modes for this bin.
+The right side of Figure~\ref{fig:chiresid} plots the individual modes
+${\rm U}_i(z),$ \ie\ the rows of the decompression matrix $\matD$ such
+that $\hat\vecn = \bar\vecn + \sum_{i\le M} u_i \vecU_i.$ Recall that
+each mode's coefficient $u_i$ will be a random deviate with unit variance.
+Each of the three modes appears to effect some combination of a $z$ shift of the main $n(z)$ peak, a change of the peak's shape/width, and a change in the low-$z$ contamination.
+We have also plotted mode 7 in the Figure, to illustrate a mode of variation that is present in the samples, but has unobservable consequences.
+Mode 7 is more oscillatory in redshift than the three retained modes, and does not alter the low-$z$ tail.
 
@@
 of the $M=3$ components of the compressed $\vecu$ representations of the
-3000 input samples.  Some are substantially non-Gaussian,
+3000 input samples, \ie\ the vectors $\vecu_j=E(\vecn_j-\bar\vecn)$ obtained from each input sample $\vecn_j$.  Some of the components of $\vecu$ have substantially non-Gaussian distributions,
 so a unit-normal prior on $\vecu$ will not faithfully represent the
@@
 Now, the cosmological Markov Chain
-is told that there are five parameters $\widetilde{\vecu}$ that have a
+is told that there are 3 parameters in $\widetilde{\vecu}$ that have a
 unit-normal prior $p(\widetilde{\vecu}).$  This vector defines the
@@
 unobservable fluctuations in $n(z)$ has substantially reduced the
-variance of the function.  To check whether we have achieved our goal
-of leaving the \emph{observable} consequences of the $n(z)$ variation
+variance of the function at any \emph{individual} $z$ value, but the \emph{collective} $n(z)$ behavior still retains the same observable influence on the summary statistics $\vecc$.
+To check whether we have achieved our goal
+of leaving the observable consequences of the $n(z)$ variation
 unchanged, we calculate the distribution of
@@
 \end{equation}
-for the cases (1) the $\vecn$ are drawn from the input samples, and
+for the cases when (1) the $\vecn$ are drawn from the input samples, vs
 (2) are generated using the procedure defined above.  This $\chi^2$
@@
   the Euclidean norm of $\vecn.$   This method enables a rigorous
-  marginalization over the distribution of $\vecn$ even in for
+  marginalization over the distribution of $\vecn$ even for
   non-Gaussian distributions, as long as the derivatives of the
@@
   \begin{acknowledgments}
-G.M.B. acknowledges support from NSF grant AST-2205808 and \ldots.
+G.M.B. acknowledges support from NSF grant AST-2205808 and DOE award DE-SC0007901.
+W. d’A. acknowledges support from the  MICINN projects PID2019-111317GB-C32, PID2022-141079NB-C32 as well as predoctoral program AGAUR-FI ajuts (2024 FI-1 00692) Joan Or\'o.
+The project that gave rise to these results received the support of a fellowship to A. Alarcon from "la Caixa" Foundation (ID 100010434). The fellowship code is LCF/BQ/PI23/11970028.
 
+Funding for the DES Projects has been provided by the U.S. Department of Energy, the U.S. National Science Foundation, the Ministry of Science and Education of Spain, 
+the Science and Technology Facilities Council of the United Kingdom, the Higher Education Funding Council for England, the National Center for Supercomputing 
+Applications at the University of Illinois at Urbana-Champaign, the Kavli Institute of Cosmological Physics at the University of Chicago, 
+the Center for Cosmology and Astro-Particle Physics at the Ohio State University,
+the Mitchell Institute for Fundamental Physics and Astronomy at Texas A\&M University, Financiadora de Estudos e Projetos, 
+Funda{\c c}{\~a}o Carlos Chagas Filho de Amparo {\`a} Pesquisa do Estado do Rio de Janeiro, Conselho Nacional de Desenvolvimento Cient{\'i}fico e Tecnol{\'o}gico and 
+the Minist{\'e}rio da Ci{\^e}ncia, Tecnologia e Inova{\c c}{\~a}o, the Deutsche Forschungsgemeinschaft and the Collaborating Institutions in the Dark Energy Survey.
+
+The Collaborating Institutions are Argonne National Laboratory, the University of California at Santa Cruz, the University of Cambridge, Centro de Investigaciones Energ{\'e}ticas, 
+Medioambientales y Tecnol{\'o}gicas-Madrid, the University of Chicago, University College London, the DES-Brazil Consortium, the University of Edinburgh, 
+the Eidgen{\"o}ssische Technische Hochschule (ETH) Z{\"u}rich, 
+Fermi National Accelerator Laboratory, the University of Illinois at Urbana-Champaign, the Institut de Ci{\`e}ncies de l'Espai (IEEC/CSIC), 
+the Institut de F{\'i}sica d'Altes Energies, Lawrence Berkeley National Laboratory, the Ludwig-Maximilians Universit{\"a}t M{\"u}nchen and the associated Excellence Cluster Universe, 
+the University of Michigan, NSF NOIRLab, the University of Nottingham, The Ohio State University, the University of Pennsylvania, the University of Portsmouth, 
+SLAC National Accelerator Laboratory, Stanford University, the University of Sussex, Texas A\&M University, and the OzDES Membership Consortium.
+
+Based in part on observations at NSF Cerro Tololo Inter-American Observatory at NSF NOIRLab (NOIRLab Prop.
+ID 2012B-0001; PI: J.
+Frieman), which is managed by the Association of Universities for Research in Astronomy (AURA) under a cooperative agreement with the National Science Foundation.
+
+The DES data management system is supported by the National Science Foundation under Grant Numbers AST-1138766 and AST-1536171.
+The DES participants from Spanish institutions are partially supported by MICINN under grants PID2021-123012, PID2021-128989 PID2022-141079, SEV-2016-0588, CEX2020-001058-M and CEX2020-001007-S, some of which include ERDF funds from the European Union.
+IFAE is partially funded by the CERCA program of the Generalitat de Catalunya.
+
+We  acknowledge support from the Brazilian Instituto Nacional de Ci\^encia
+e Tecnologia (INCT) do e-Universo (CNPq grant 465376/2014-2).
+
+This document was prepared by the DES Collaboration using the resources of the Fermi National Accelerator Laboratory (Fermilab), a U.S.
+Department of Energy, Office of Science, Office of High Energy Physics HEP User Facility.
+Fermilab is managed by Fermi Forward Discovery Group, LLC, acting under Contract No.
+89243024CSC000002.
 \end{acknowledgments}
+
+\textbf{Contributions:}
+GB devised the compression method, wrote the relevant code, and wrote the article text.
+MT derived the relevant derivative matrices, and Wd'A applied the method to the DES Maglim bin and produced figures.
+Authors Wd'A, MT, AA, AA, GG, and BY developed the redshift samples used in the demonstration, integrated the code into DES pipelines, and advised on the implementation and text.
+The remaining authors have made contributions to this paper that include, but
+are not limited to, the construction of DECam and other aspects of collecting the data; data
+processing and calibration; developing broadly used methods, codes, and simulations; running
+the pipelines and validation tests; and promoting the science analysis.
+
 \begin{thebibliography}{dummy}
@@
 %\bibliographystyle{aasjournal}
-\bibitem[d'Assignies et al.(2025)]{y6wz} d'Assignies, W. et al.\ 2025 (WZ paper)
+\bibitem[d'Assignies et al.(2025)]{y6wz} d'Assignies, W. et al.\ 2025
+  (WZ paper) ???
+\bibitem[Bridle et al.(2002)]{bridle02}
+Bridle, S. L., Crittenden, R., Melchiorri, A., Hobson, M. P., Kneissl, R., and Lasenby, A. N. 2002, \mnras, 335, 1193, 
+doi:10.1046/j.1365-8711.2002.05709.x
 
@@
 
-\bibitem[Giannini et al.(2025)]{y6gg} Giannini, G. et al.\ 2025 (lenses paper)
+\bibitem[Giannini et al.(2025)]{y6gg} Giannini, G. et al.\ 2025
+  (lenses paper) ????
 
@@
 
-\bibitem[Yin et al.(2025)]{y6pz} Yin, B. et al.\ 2025 (PZ paper)
+\bibitem[Yin et al.(2025)]{y6pz} Yin, B. et al.\ 2025 (PZ paper) ???
 

### 2025-05-23 authors.tex
post-CWR updates including author list, aastex7

% Author list file generated with: mkauthlist 1.3.0+30.g918973f.dirty 
% mkauthlist -j aastex7 -sb --orcid author_list.csv

\suppressAffiliations
\correspondingauthor {G.~M.~Bernstein}

\author[0000-0002-8613-8259, gname='Gary', sname='Bernstein']{G.~M.~Bernstein}
\affiliation{Department of Physics and Astronomy, University of Pennsylvania, Philadelphia, PA 19104, USA}
\email{garyb@physics.upenn.edu}

\author[0000-0002-9719-1717, gname='William', sname='Assignies Doumerg']{W.~d'Assignies D.}
\affiliation{Institut de F\'{\i}sica d'Altes Energies (IFAE), The Barcelona Institute of Science and Technology, Campus UAB, 08193 Bellaterra (Barcelona) Spain}
\email{wdoumerg@igfae.es}

\author[0000-0002-5622-5212, gname='Michael A.', sname='Troxel']{M.~A.~Troxel}
\affiliation{Department of Physics, Duke University Durham, NC 27708, USA}
\email{michael.a.troxel@gmail.com}

\author[0000-0001-8505-1269, gname='Alex', sname='Alarcon']{A.~Alarcon}
\affiliation{Institute of Space Sciences (ICE, CSIC),  Campus UAB, Carrer de Can Magrans, s/n,  08193 Barcelona, Spain}
\email{alexalarcongonzalez@gmail.com}

\author[0000-0002-6445-0559, gname='Alexandra', sname='Amon']{A.~Amon}
\affiliation{Department of Astrophysical Sciences, Princeton University, Peyton Hall, Princeton, NJ 08544, USA}
\email{alexandra.amon@princeton.edu}

\author[0000-0002-3730-1750, gname='Giulia', sname='Giannini']{G.~Giannini}
\affiliation{Kavli Institute for Cosmological Physics, University of Chicago, Chicago, IL 60637, USA}
\email{giulia.giannini15@gmail.com}

\author[0000-0001-7005-8820, gname='Boyan', sname='Yin']{B.~Yin}
\affiliation{Department of Physics, Carnegie Mellon University, Pittsburgh, Pennsylvania 15312, USA}
\email{byin@andrew.cmu.edu}

\author[0000-0002-7069-7857, gname=Sahar, sname=Allam]{S.~S.~Allam}
\affiliation{Fermi National Accelerator Laboratory, P. O. Box 500, Batavia, IL 60510, USA}
\email{sallamatfnalgov@gmail.com}

\author[0000-0003-0171-6900, gname='Felipe', sname='Andrade-Oliveira']{F.~Andrade-Oliveira}
\affiliation{Physik-Institut, University of Zürich, Winterthurerstrasse 190, CH-8057 Zürich, Switzerland}
\email{felipeaoli@gmail.com}

\author[0000-0002-8458-5047, gname='David', sname='Brooks']{D.~Brooks}
\affiliation{Department of Physics \& Astronomy, University College London, Gower Street, London, WC1E 6BT, UK}
\email{david.brooks@ucl.ac.uk}

\author[0000-0003-3044-5150, gname='Aurelio', sname='Carnero Rosell']{A.~Carnero~Rosell}
\affiliation{Instituto de Astrofisica de Canarias, E-38205 La Laguna, Tenerife, Spain}
\affiliation{Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Av. Pastor Martin Luther King Jr, 126 Del Castilho, Nova Am\'erica Offices, Torre 3000/sala 817 CEP: 20765-000, Brazil}
\affiliation{Universidad de La Laguna, Dpto. Astrofísica, E-38206 La Laguna, Tenerife, Spain}
\email{aurelio.crosell@gmail.com}

\author[0000-0002-3130-0204, gname='Jorge', sname='Carretero']{J.~Carretero}
\affiliation{Institut de F\'{\i}sica d'Altes Energies (IFAE), The Barcelona Institute of Science and Technology, Campus UAB, 08193 Bellaterra (Barcelona) Spain}
\email{jorgecarreteropalacios@gmail.com}

\author[0000-0002-7731-277X, gname='Luiz', sname='da Costa']{L.~N.~da Costa}
\affiliation{Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Av. Pastor Martin Luther King Jr, 126 Del Castilho, Nova Am\'erica Offices, Torre 3000/sala 817 CEP: 20765-000, Brazil}
\email{ldacosta@fnal.gov}

\author[0000-0000-0000-0000,gname='Maria Elidaiana', sname='da Silva Pereira']{M.~E.~S.~Pereira}
\affiliation{Hamburger Sternwarte, Universit\"{a}t Hamburg, Gojenbergsweg 112, 21029 Hamburg, Germany}
\email{elidaiana.sp@gmail.com}

\author[0000-0001-8318-6813, gname='Juan', sname='De Vicente']{J.~De~Vicente}
\affiliation{Centro de Investigaciones Energ\'eticas, Medioambientales y Tecnol\'ogicas (CIEMAT), Madrid, Spain}
\email{juan.vicente@ciemat.es}

\author[0000-0000-0000-0000,gname='Spencer', sname='Everett']{S.~Everett}
\affiliation{California Institute of Technology, 1200 East California Blvd, MC 249-17, Pasadena, CA 91125, USA}
\email{spencer.w.everett@jpl.nasa.gov}

\author[0000-0003-4079-3263, gname='Josh', sname='Frieman']{J.~Frieman}
\affiliation{Department of Astronomy and Astrophysics, University of Chicago, Chicago, IL 60637, USA}
\affiliation{Fermi National Accelerator Laboratory, P. O. Box 500, Batavia, IL 60510, USA}
\affiliation{Kavli Institute for Cosmological Physics, University of Chicago, Chicago, IL 60637, USA}
\email{frieman@fnal.gov}

\author[0000-0002-9370-8360, gname='Juan', sname='Garcia-Bellido']{J.~Garc\'ia-Bellido}
\affiliation{Instituto de Fisica Teorica UAM/CSIC, Universidad Autonoma de Madrid, 28049 Madrid, Spain}
\email{juan.garciabellido@uam.es}

\author[0000-0003-3270-7644, gname='Daniel', sname='Gruen']{D.~Gruen}
\affiliation{University Observatory, Faculty of Physics, Ludwig-Maximilians-Universit\"at, Scheinerstr. 1, 81679 Munich, Germany}
\email{dgruen@usm.uni-muenchen.de}

\author[0000-0003-2071-9349, gname='Samuel', sname='Hinton']{S.~R.~Hinton}
\affiliation{School of Mathematics and Physics, University of Queensland,  Brisbane, QLD 4072, Australia}
\email{samuelreay@gmail.com}

\author[0000-0002-9369-4157, gname='Devon L.', sname='Hollowood']{D.~L.~Hollowood}
\affiliation{Santa Cruz Institute for Particle Physics, Santa Cruz, CA 95064, USA}
\email{dhollowo@ucsc.edu}

\author[0000-0002-6550-2023, gname='Klaus', sname='Honscheid']{K.~Honscheid}
\affiliation{Center for Cosmology and Astro-Particle Physics, The Ohio State University, Columbus, OH 43210, USA}
\affiliation{Department of Physics, The Ohio State University, Columbus, OH 43210, USA}
\email{kh@fnal.gov}

\author[0000-0001-5160-4486, gname='David', sname='James']{D.~J.~James}
\affiliation{Center for Astrophysics $\vert$ Harvard \& Smithsonian, 60 Garden Street, Cambridge, MA 02138, USA}
\email{djames44@gmail.com}

\author[0000-0002-8289-740X, gname='Sujeong', sname='Lee']{S.~Lee}
\affiliation{Jet Propulsion Laboratory, California Institute of Technology, 4800 Oak Grove Dr., Pasadena, CA 91109, USA}
\email{sujeong.lee@jpl.nasa.gov}

\author[0000-0003-0710-9474, gname='Jennifer', sname='Marshall']{J.~L.~Marshall}
\affiliation{George P. and Cynthia Woods Mitchell Institute for Fundamental Physics and Astronomy, and Department of Physics and Astronomy, Texas A\&M University, College Station, TX 77843,  USA}
\email{jlm076@tamu.edu}

\author[0000-0001-9497-7266, gname='Juan', sname='Mena-Fernández']{J. Mena-Fern{\'a}ndez}
\affiliation{Universit\'e Grenoble Alpes, CNRS, LPSC-IN2P3, 38000 Grenoble, France}
\email{juan.menafernandez@lpsc.in2p3.fr}

\author[0000-0002-6610-4836, gname='Ramon', sname='Miquel']{R.~Miquel}
\affiliation{Instituci\'o Catalana de Recerca i Estudis Avan\c{c}ats, E-08010 Barcelona, Spain}
\affiliation{Institut de F\'{\i}sica d'Altes Energies (IFAE), The Barcelona Institute of Science and Technology, Campus UAB, 08193 Bellaterra (Barcelona) Spain}
\email{rmiquel@ifae.es}

\author[0000-0002-2598-0514, gname='Andrés', sname='Plazas Malagón']{A.~A.~Plazas~Malag\'on}
\affiliation{Kavli Institute for Particle Astrophysics \& Cosmology, P. O. Box 2450, Stanford University, Stanford, CA 94305, USA}
\affiliation{SLAC National Accelerator Laboratory, Menlo Park, CA 94025, USA}
\email{plazasmalagon@gmail.com}

\author[0000-0002-9646-8198, gname='Eusebio', sname='Sanchez']{E.~Sanchez}
\affiliation{Centro de Investigaciones Energ\'eticas, Medioambientales y Tecnol\'ogicas (CIEMAT), Madrid, Spain}
\email{eusebio.sanchez@ciemat.es}

\author[0000-0003-3054-7907, gname='David', sname='Sanchez Cid']{D.~Sanchez Cid}
\affiliation{Centro de Investigaciones Energ\'eticas, Medioambientales y Tecnol\'ogicas (CIEMAT), Madrid, Spain}
\affiliation{Physik-Institut, University of Zürich, Winterthurerstrasse 190, CH-8057 Zürich, Switzerland}
\email{david.sanchez@ciemat.es}

\author[0000-0002-1831-1953, gname='Ignacio', sname='Sevilla']{I.~Sevilla-Noarbe}
\affiliation{Centro de Investigaciones Energ\'eticas, Medioambientales y Tecnol\'ogicas (CIEMAT), Madrid, Spain}
\email{nsevilla@fnal.gov}

\author[0000-0000-0000-0000,gname='Tae-hyeon', sname='Shin']{T.~Shin}
\affiliation{Department of Physics and Astronomy, Stony Brook University, Stony Brook, NY 11794, USA}
\email{babohahaman@gmail.com}

\author[0000-0002-3321-1432, gname='Mathew', sname='Smith']{M.~Smith}
\affiliation{Physics Department, Lancaster University, Lancaster, LA1 4YB, UK}
\email{mat.smith@soton.ac.uk}

\author[0000-0002-7047-9358, gname='Eric', sname='Suchyta']{E.~Suchyta}
\affiliation{Computer Science and Mathematics Division, Oak Ridge National Laboratory, Oak Ridge, TN 37831}
\email{eric.d.suchyta@gmail.com}

\author[0000-0000-0000-0000,gname='Molly', sname='Swanson']{M.~E.~C.~Swanson}
\affiliation{Center for Astrophysical Surveys, National Center for Supercomputing Applications, 1205 West Clark St., Urbana, IL 61801, USA}
\email{molly.swanson@gmail.com}

\author[0000-0001-9382-5199, gname='Noah', sname='Weaverdyck']{N.~Weaverdyck}
\affiliation{Department of Astronomy, University of California, Berkeley,  501 Campbell Hall, Berkeley, CA 94720, USA}
\affiliation{Lawrence Berkeley National Laboratory, 1 Cyclotron Road, Berkeley, CA 94720, USA}
\email{nweaverd@umich.edu}

\author[0000-0002-8282-2010, gname='Jochen', sname='Weller']{J.~Weller}
\affiliation{Max Planck Institute for Extraterrestrial Physics, Giessenbachstrasse, 85748 Garching, Germany}
\affiliation{Universit\"ats-Sternwarte, Fakult\"at f\"ur Physik, Ludwig-Maximilians Universit\"at M\"unchen, Scheinerstr. 1, 81679 M\"unchen, Germany}
\email{jochen.weller@lmu.de}

\author[0000-0002-3073-1512, gname='Philip', sname='Wiseman']{P.~Wiseman}
\affiliation{School of Physics and Astronomy, University of Southampton,  Southampton, SO17 1BJ, UK}
\email{P.S.Wiseman@soton.ac.uk}

% Number in brackets is how many author names to put on front page
\collaboration{7}{(DES Collaboration)}

### 2025-05-23 main.tex
post-CWR updates including author list, aastex7

@@
 % \documentclass[preprint,linenumbers]{aastex631} # 6.31 fails with \align!!
-\documentclass[linenumbers, onecolumn]{aastex63}
+\documentclass[linenumbers, onecolumn]{aastex7}
 \usepackage{xcolor}
@@
 
-\author[0000-0002-8613-8259]{Gary M. Bernstein}
-\affiliation{Department of Physics and Astronomy, University of Pennsylvania, Philadelphia, PA 19104, USA}
-\email{garyb@upenn.edu}
+\include{authors.tex}
+% \author[0000-0002-8613-8259]{Gary M. Bernstein}
+% \affiliation{Department of Physics and Astronomy, University of Pennsylvania, Philadelphia, PA 19104, USA}
+% \email{garyb@upenn.edu}
 
-\author{William d'Assignies}
-\author{Michael Troxel}
-\author{Alex Alarcon}
-\author{Alex Amon}
-\author{Giulia Giannini}
-\author{Boyan Yin}
-\author{Builders\ldots} %%%%
+% \author{William d'Assignies}
+% \author{Michael Troxel}
+% \author{Alex Alarcon}
+% \author{Alex Amon}
+% \author{Giulia Giannini}
+% \author{Boyan Yin}
+% \author{Builders\ldots} %%%%
 
@@
 A typical Bayesian inference on the values of some parameters of
-interest $\vecq$ from some data $D$ would run a Markov Chain (MC) to sample
+interest $\vecq$ from some data $D$ involves running a Markov Chain (MC) to sample
 from the posterior $p(\vecq,\vecn | D) \propto \likeli(D | \vecq,\vecn)
@@
   itself known only from a set of samples of $\vecn$ from this
-  distribution.} Most MC samplers require that the posterior be an
+  distribution.} Most MC samplers require that the posterior (and
+hence the prior and likelihood) be an
 evaluable function of any value of the parameters.
@@
 colors in an attempt to generate a sample that is confined to a
-particular redshift range, as described in \citet{y6gg}.  Once each bin's
+particular redshift range, as described in \citet{y6maglim}.  Once each bin's
 sample is chosen, a combination of 
-photometric techniques \citep{y6pz} and clustering information
+photometric techniques \citep{y6lenspz,y6pz} and clustering information
 \citep{y6wz} is used to generate samples from the posterior
@@
 
+For simplicity, we have demonstrated this method for a case in which
+$\vecn$ specifies the $n(z)$ function for a single sample of DES
+galaxies.  The method is fully applicable to any set of nuisance
+parameters for which we are given a set of samples from $p(\vecn).$
+For instance, in the DES Y6 analysis, we use a $\vecn$ that is a
+concatenation of the parameters of $n(z)$ for 4 bins of lens source
+galaxies.  Since each output sample from the  redshift-estimate
+process describes all 4 bins, we know the cross-correlations between
+the bins' $n(z)$ behavior, and the compression process described
+herein will properly preserve these correlations in the subsequent
+cosmological analysis.
+
 \section{Summary}
@@
 %\bibliographystyle{aasjournal}
+% Y6 WZ paper
 \bibitem[d'Assignies et al.(2025)]{y6wz} d'Assignies, W. et al.\ 2025
-  (WZ paper) ???
+  (in preparation)
+
 \bibitem[Bridle et al.(2002)]{bridle02}
@@
 
-\bibitem[Giannini et al.(2025)]{y6gg} Giannini, G. et al.\ 2025
-  (lenses paper) ????
+% lenses paper
+\bibitem[Giannini et al.(2025)]{y6lenspz} Giannini, G. et al.\ 2025 (in preparation)
 
@@
 
-\bibitem[Yin et al.(2025)]{y6pz} Yin, B. et al.\ 2025 (PZ paper) ???
+\bibitem[Giannini et al.(2025)]{y6maglim} Weaverdyck, N. et al.\ 2025 (in preparation)
+
+\bibitem[Yin et al.(2025)]{y6pz} Yin, B. et al.\ 2025 (in preparation)
 
@@
 
-\bibitem[??? et al.(2025)]{y6model}
-  Y6 modeling paper ???
+% Y6 modelling paper
+\bibitem[Sanchez-Cid et al.(2025)]{y6model} Sanchez-Cid et al.\ 2025
+  (in preparation)
 
@@
 
+\allauthors
 \end{document}

### 2025-05-25 main.tex
Ready for final reading

@@
 \shortauthors{Bernstein et al.}
+\shorttitle{Sampling sampled priors}
+\reportnum{DES-2025-0886}
+\reportnum{FERMILAB-PUB-25-0362-PPD}
 
@@
 \newcommand{\trace}{\text{Tr}}
+
 \begin{document}
@@
 \include{authors.tex}
-% \author[0000-0002-8613-8259]{Gary M. Bernstein}
-% \affiliation{Department of Physics and Astronomy, University of Pennsylvania, Philadelphia, PA 19104, USA}
-% \email{garyb@upenn.edu}
-
-% \author{William d'Assignies}
-% \author{Michael Troxel}
-% \author{Alex Alarcon}
-% \author{Alex Amon}
-% \author{Giulia Giannini}
-% \author{Boyan Yin}
-% \author{Builders\ldots} %%%%
+\correspondingauthor{G.~M.~Bernstein}
 
@@
 \end{abstract}
-\reportnum{}
 
@@
 advantage of our method over the single-step covariance-inflation
-method of \citet{hans}.
-But even if the \vecu\ prior is well approximated as multivariate normal,
+method of \citet{hans}.  A further advantage of our
+compression$+$density-estimator scheme over covariance inflation is
+that nonlinearities in the data model, including presence of
+significant $\partial^2\hatc/\partial\vecq\partial\vecn,$ will be
+correctly handled in the inference to the extent that the higher
+derivatives are confined to the compressed subspace of $\vecn.$ 
+But even in the case where linearity and gaussianity in \vecu\ hold,
 there are practical advantages of compressing the nuisance 
@@
 herein will properly preserve these correlations in the subsequent
-cosmological analysis.
+cosmological analysis.  In the DES Y6 analysis, we have also extended
+the nuisance parameter vector $\vecn$ to include the multiplicative
+errors on the shear measurement method, and create a compressed
+$\vecu$ that captures correlations between the multiplicative errors
+and the $n(z)$ estimates.
 
@@
   log-likelihood of the data with respect to $\vecn$ are nearly
-  constant over the posterior domain.  The mode-projection technique
+  constant over the posterior domain.  Indeed the effect of the
+  nuisance parameters on the inference will be correctly calculated
+  even if these derivatives vary across the posterior domain, as long
+  as the statistically relevant values of $\vecn$ do not deviate
+  significantly from the linear subspace defined by our compression.
+The mode-projection technique
   is also useful as a method of projecting away irrelevant

### 2025-06-01 main.tex
arxiv and apj submission

@@
 % \documentclass[preprint,linenumbers]{aastex631} # 6.31 fails with \align!!
-\documentclass[linenumbers, onecolumn]{aastex7}
+%\documentclass[linenumbers, onecolumn]{aastex7}
+\documentclass[onecolumn]{aastex7}
 \usepackage{xcolor}
@@
 of 2-point correlation functions of weak lensing fields and galaxy
-density in the \textit{Dark Energy Survey}, where $\vecn$ is a binned representation of the redshift distrbution
+density in the \textit{Dark Energy Survey}, where $\vecn$ is a binned representation of the redshift distribution
 $n(z)$ of the galaxies.
@@
 \citep{y6wz} is used to generate samples from the posterior
-distribtuion of  the $n(z)$ distribution of the bin members, conditioned
+distribution of  the $n(z)$ distribution of the bin members, conditioned
 on the photometric and clustering data.
@@
 parameters for which we are given a set of samples from $p(\vecn).$
-For instance, in the DES Y6 analysis, we use a $\vecn$ that is a
+For instance, in the DES Y6 analysis, we use an $\vecn$ that is a
 concatenation of the parameters of $n(z)$ for 4 bins of lens source

### 2025-08-22 response.txt
Resubmission version

> 
> The authors present an intriguing linear compression and decompression technique
> for Bayesian inference when the prior distribution of nuisance parameters is
> available only through samples from prior research. The mathematical concepts
> are engaging, and the method is illustrated with a numerical example. However, I
> have some suggestions to enhance clarity, rigor, and comparability with existing
> methods. Below, I outline my comments to support the authors in refining their
> work.
> 
> 1. Title: The phrase "Sampling Bayesian probabilities" may be
> unclear. Probability, as a mapping from an event to a value between 0 and 1, is
> consistent across Bayesian and frequentist frameworks. Could the authors clarify
> what is meant by "Bayesian probabilities" in the main text and consider a title
> that more precisely reflects the manuscript's focus? For example, if the intent
> is to generate samples from a posterior distribution, a phrase like "sampling
> from a probability distribution" might be more standard. A revised title could
> better capture the core contribution.
>

We have a new title, "Sampling posterior distributions when only samples from a prior are available"

> 2. Abstract: The assumption that the joint prior p(q, n) factorizes as p(q)p(n)
> holds only when q and n are independent. As the abstract discusses Bayesian
> inference generally, it may be helpful to clarify this assumption or use a more
> general expression for the joint prior.

It is now clear in the abstract and text that we are assuming separable priors for the nuisance parameters and the parameters of interest (Line 26).

> 
> 3. Abstract: The claim that the issue of a prior p(n) defined only by samples is
> common may benefit from adjustment. If this is a specific case, replacing "many"
> with "some" could enhance clarity. If the authors believe this issue is
> widespread, citing Bayesian textbooks or references addressing this scenario
> would strengthen the claim. Additionally, to ensure adherence to Bayesian
> principles, could the authors confirm that the data used in this study are
> distinct from those used to construct the prior samples, avoiding potential
> reuse of data?
>

"many" -> "some" in the abstract.  And in the opening paragraph, we add text stating that the prior on nuisance parameters would come from distinct observational data (line 29).

> 4. Abstract and Section 1: The likelihood function, as defined by R. A. Fisher,
> is typically a function of model parameters given the data, denoted as L(q, n|D)
> or L(q, n). The current notation L(D | q, n) as a function of data given
> parameters may cause confusion. If the authors intend to use f(D|q, n) to
> represent the probability density of the data given parameters, explicitly
> defining it as L(q, n) = f(D|q, n) could clarify this. A reference to standard
> definitions, such as those on Wikipedia, might help.
> 
> 5. Section 1: The term "Bayesian posterior probability" before Equation (1)
> might be more accurately described as the "Bayesian posterior density," as
> Equation (1) represents a density, not a probability.
>

Our nomenclature is in agreement already with that given in the Wikipedia entry for "Posterior probability", which states Bayes's Thm as "The posterior probability is therefore proportional to the product Likelihood · Prior probability," citing Ross.  We have placed a footnote to that effect on the opening page.

> 6. Section 1: The statement that the prior p(n) is known only through samples
> could be clarified. If the authors assume these samples are drawn from an
> unknown prior p(n), revising the sentence to avoid implying knowledge of p(n)
> might reduce confusion. For example, stating that the samples are assumed to
> represent p(n) could be clearer.
>

There is now explicit text that we consider a case where p(n) cannot be directly evaluated (line 30).

> 7. Section 1: The manuscript notes that insufficient samples for
> high-dimensional n may hinder density estimation but does not appear to address
> this issue. Could the authors clarify their proposed solution for cases with
> limited samples, or confirm if this remains a challenge under the
> compression-decompression approach?
>

We believe that the projection of the prior into a space of much smaller dimension *is* the proposed solution to this problem, since far fewer samples are necessary to build a usable density estimator in the compressed space.  In our example case of an 80-dimensional nuisance vector, density estimation was impossible in the full space even with millions of samples, but thousands of samples more than suffice to build the density estimator in the 3-dimensional compressed space.

> 8. Section 1: The discussion of running a new Markov Chain (MC) for each sample
> of n to marginalize over n is described as infeasible for a large number of
> samples. Could the authors elaborate on why this is infeasible and whether a
> smaller number of samples would make it feasible? Additionally, specifying what
> constitutes a "large" or "small" number of samples (e.g., 100 or 1000 vectors)
> would provide useful context, especially given the numerical example.
>

We now state "thousands or more" instand of "large number" - although in many of our chains, running even one full chain is considered expensive, so having to repeat this even 10 times could cross the threshold out of affordability (line 48).

> 9. Section 1: The notation c_hat for the model is unconventional, as a hat
> typically denotes an estimator. Consider using c = g(q, n) for a deterministic
> model or g(c|q, n) for a probabilistic one to align with standard notation.
>
We've changed the notation for the (noiseless) data to \hat c, since this will be the mean of the model normal distribution for the observed (noisy) data c.

> 10. Section 1: The notation b_k(z) suggests a deterministic function of redshift
> z. Providing its functional form, perhaps in an appendix, would enhance clarity,
> given its relevance to prior construction.
>

b_k is indeed a deterministic function.  We now have a clause giving an example of what these redshift "kernels" could be (line 41).

> 11. Section 1: The manuscript could explore alternative Bayesian approaches,
> such as sampling the full posterior p(q, n|c) using a Gibbs-type algorithm or
> Hamiltonian Monte Carlo, leveraging the linearity and Gaussian
> assumptions. Additionally, since p(n) is unavailable, using an estimated density
> p_hat(n) might be viable. For the Gaussian case, analytically marginalizing n to
> obtain p(c|q) seems feasible, as it involves computing <n> and C_n from the
> samples of n (e.g., 3000 samples of n in the example). Could the authors discuss
> why this analytic approach is not pursued, particularly for Gaussian priors, and
> clarify the role of continuity in p(u) if n is marginalized in such an analytic
> manner?
>

As the paper now more clearly states, we are considering a case where an evaluable p(n) is not available. So analytic marginalization is not going to be an option.  And as is stated now near lines 33, a density estimator over the full n space is often infeasible, so this is the case we are trying to address - constructing a viable density estimator, but on a space of reduced dimension.  HMC is an example of an algorithm that requires an evaluable p(n) (and a differentiable one, in addition), and likewise Gibbs sampling requires some kind of sampling method over subsets of the full parameter space, leaving us back where we started, needing to sample from p(n).  So we feel that our statement that p(n) is not evaluable, and most MC methods will require an evaluable p(n), is already excluding the alternative methods that the referee suggests we discuss.

> 12. Section 1: The term "MC chain" may be redundant if MC stands for Markov
> Chain. Consider using "Markov Chain" or "MC" consistently.
>

I'm embarrassed to have used "MC chain" because I constantly tell other people that it's redundant :)
These cases have been fixed.

> 13. Section 1: To ensure the validity of the Gaussian prior in Step 2, could the
> authors confirm that the n samples are independent of the data used for the
> summary statistic c?
>

This is now made explicit in the opening paragraph.

> 14. Section 1: The term n_alpha is unclear. Is alpha an index or a marker for
> compressed/decompressed samples? Later, alpha appears as a summation index in
> Equation (3). Clarifying its usage would help.
>

We have now moved into Section 1, near line 51, the exposition that \alpha is an index over the given samples of the n vector.
The mean is hence over all of the samples of the nuisance parameters, and this is its meaning in the summations as well.

> 15. Section 1: The statement that the method maintains "equal prior probability
> of each n sample" is unclear. Could the authors explain whether this means
> preserving the prior distribution post-compression-decompression?
>
Line 54 now has "gives each sample $\vecn_\alpha$ equal probability under
  $p(\vecu),$ maintaining the meaning of the input samples"
which we hope is clear.

> 16. Section 1: The claim that the likelihood function is discontinuous is
> introduced late. Could the authors specify which parameters cause this
> discontinuity and provide details, perhaps in an appendix, when introducing the
> likelihood function?

The text is we hope now clearer, that the discontinuity arises in the mapping from the u coordinates of the hypercube into the n values that have been placed into the grid embedded in the hypercube.  The nearest-neighbor assignment is discontinuous (Line 54).

> 
> 17. Section 1: The assertion that marginalizing n equates to adding terms to Cc
> is not immediately clear. This seems to hold only if the coefficient of n in
> c_hat(q, n) is 1. Providing the functional form of this linear function would
> clarify this.
>

To keep our paper succinct, we prefer not to give the derivation of a method that we will not be using.  The text now just says that this equivalence is shown by the cited authors.

> 18. Section 1: The claim that the method is a good approximation when conditions
> are not grossly violated requires justification. Could the authors provide
> reasoning or references to support this?
>
We have removed that clause.  There is a sentence near the end of section 2 that explains a bit more, namely that the compression scheme we develop will still be valid if 1 & 3 are violated but the compression still retains nearly all the cosmologically relevant information in n (Line 175)

> 19. Section 1: Specifying the typical dimensionality of n in practice (e.g.,
> before the numerical example's 800 dimensions) would provide valuable context.

The dimensionality of the example application is now mentioned at the end of Section 1 (Line 78).

> 
> 20. Section 1: In Step 3, the linearization of c_hat around fiducial values q0,
> n0 is unclear. Could the authors provide mathematical details, perhaps in an
> appendix, and explain why derivatives are independent of q? Background
> supporting the claimed accuracy relative to measurement errors would also help.
>

This list is of restrictions, so the linearization is a requirement, not something that is derived.

> 21. Section 1: The Gaussian assumption for data and priors may not always
> hold. Could the authors discuss the robustness of this assumption, perhaps
> considering Student's t distributions for data with outliers? Additionally,
> clarifying the scientific benefit of accommodating non-Gaussian priors on n with
> a numerical example would strengthen the case.
>

The Gaussian assumption for the likelihood (condition 1) is essential the derivation, and we consider it beyond the scope of this paper to address other possible likelihood functions.  The Gaussian assumption for p(n) (condition 2) is one that we do not need to make for our method.

> 22. Section 2: The mean adjustment of n vectors by n0 = ⟨nα⟩ seems to precede
> the availability of n_alpha, which results from decompression. Could the authors
> clarify this sequence?
>

We hope the text is now clearer that n_\alpha are the input to our process, but we are going to subtract their collective mean before the subsequent algebraic manipulations (Line 82).

> 23. Section 2: Minimizing the Gaussian mean shift in c does not clearly ensure
> minimal impact on the posterior distribution. Could the authors explain how this
> guarantees accurate posterior inference or whether the posterior remains
> unaffected if the mean shift is zero?

Do we understand correctly that  the referee is asking why the minimization of the quantity in Eqn 3 is achieving the stated goal of finding the best compression?  The sentences following this equation indicate that minimization of this quantity is indeed equivalent to the best possible reproduction of the likelihood only in a special case.  So there is no formal proof (that we have found) that our procedure is generally optimal.  We have added text at line 93 that, we hope, makes it clear that the <chi^2> quantity is a useful proxy to minimize, since attaining zero value would imply that "no probabilities have been harmed" by the compression.

> 
> 24. Section 2: The notation n and n_alpha is confusing, as n_alpha is described
> as a decompression result yet undergoes further compression-decompression
> (n_alpha -> u_alpha -> n_hat_alpha). Could the authors clarify or correct this?
>

We hope this is clearer after having stated explicitly that alpha is an index into the input samples of n.  Then the n_alpha -> u_alpha -> n_hat_alpha notation indicates that each individual sample is compressed and then decompressed.

> 25. Section 2: The choice of dimension M for u relative to N lacks rigor. Could
> the authors provide a mathematical justification for selecting M, perhaps
> comparing to PCA's elbow rule?
>

The choice of M is indeed not rigorous, it's a heuristic like the "elbow rule."  But unlike generic PCAs, the chi^2 values do have an absolute scale attached since they are logs of probabilities.  New text in step 3 of the procedure gives a bit more incentive for the (crude) cutoff criterion as being related to small errors in the posterior values that our algorithm is helping to calculate (Line 162)

> 26. Section 2: The goal of minimizing cosmological inference errors is noted,
> but the optimization minimizes mean shifts in the likelihood, not the
> posterior. Could the authors justify this or consider optimizing the posterior
> distribution distance?
>

This distinction is indeed correct, but we do not have a solution for formally optimizing the errors in posterior values or any kind of metric for the distribution.  As we do note (Line 91ff), the optimization criterion for the compression scheme is not really optimizing any formal quantity on the posterior.  We now include the cruder statement that, in essence, getting the likelihood correct to <1% means that one also has the posterior correct to this level (since the latter is an integral over the former times a non-negative prior).

> 27. Section 2: The covariance matrix C_c is described as given. Could the
> authors clarify how it is defined or computed?
>

It has the standard meaning of covariance of experimental realizations of c at fixed values of q,n.  It is in the definition of the first required condition in Section 1.  Our algorithm is independent of how it has been obtained, whether through analytic calculations, jackknifing of the data, or some combination, so we opted to not enumerate the possibilities.

> 28. Summary Steps (Pages 4-5): The steps appear informal. Presenting them in an
> algorithmic format, specifying inputs and iteration details for sampling q and
> u, would enhance reproducibility.
>

We hope that narrative statements are clear for this description, since each step cites the equation(s) to be used to execute the step, making them formally well defined.  The except is step 6 - constructing a density estimator in the compressed space - where the user could choose any of the many available methods.

> 29. Section 3: The phrase "posterior distribution of the n(z) distribution" is
> unclear. Could the authors provide a mathematical definition?

New text at line 190:  generate samples from the posterior
distribution of  $n(z)$ functions applicable to the bin members, conditioned
on the photometric and clustering data, \ie\ the $p(n)$ that will become the
  prior for the inference of \vecq.

> 
> 30. Section 3: The term "detectable levels" in altering c_hat(n) is vague. Could
> the authors specify what levels are meant?

New text at line 201: "levels comparable to the
  precision of the observations."

> 
> 31. Section 3: The legend in the right panel of Figure 2 appears
> incomplete. Could this be corrected?
>

There are only two cases shown in this corner plot, the first described by the orange-y lines and shading, the second by the blue lines and contours.

> 32. Section 3: Comparing the proposed method with existing approaches (e.g.,
> Cordero et al., Bridle et al., Hadzhiyska et al.) would strengthen the
> manuscript. A numerical example highlighting the method's advantages over
> competitors could demonstrate its value. Additionally, the single example may
> not convince readers of general applicability. Could the authors provide
> evidence of robustness across datasets?
>

We have noted in the introduction that the Cordero et al method was not able to produce any usable results for the example problem.  We agree that the paper would be stronger with direct comparison to the Bridle method; however the effort to do so would be very large, because the process of producing cosmological constraints is very long and involved -  there will be *many* pages of journal articles forthcoming soon about this. Reproducing this process with alternative methods is an unaffordable effort.  We do now refer the readers to the papers (in prep) where this method is used extensively, with some measures of how the compression process has (not) altered inferences.

> 33. Section 3: Fully specifying the target posterior f(q, n|c or D)
> mathematically, perhaps in an appendix, would aid reproducibility.

If the referee is referring to describing the posterior for the cosmological parameters in the DES analyses: this is also quite involved, too much so to fit into an Appendix.  A (long) paper in preparation describes the analytical and numerical model for the c quantities in the DES Y6 analyses - see https://arxiv.org/abs/2105.13548 for the modeling of the 3-year data if interested.

> 
> 34. Section 3: The threshold "chisq < 0.025" seems arbitrary. Could the authors
> define accuracy in units comparable to measurement error to justify this choice,
> as the chisq quantity is unitless?
>

The chisq values are already in units of statistical significance, i.e. scaled by the measurement errors as embodied in C_c.  We added a few words at line 215 describing our motivation - to keep the total impact of compressing 10 modes to be well below the 1-sigma uncertainties inherent to the data.  There is not a precise motivation.

> 35. Section 3: The CDF normalization procedure's impact on preserving the target
> distribution is unclear. Could the authors provide references or proofs
> supporting its validity?
>

The implicit function using CDFs in Eqn. 23 is the standard variable-transformation method for sampling an arbitrary 1d probability distribution, so we hope it does not require a proof here.   As to what they do, we refer to the paragraph starting at line 222, which notes that the histogram for, in particular, the u1  histogram in Figure 3, is by eyeball inspection definitely non-Gaussian.  So to sample from p(u) [and therefore sample from p(n)] we need to be able to sample from this histogram.  Hence the use of the variable-transformation to convert a non-Gaussian distribution into a Gaussian one.

> 36. Section 3: Using a chi-square measure for Bayesian inference is
> unconventional. Consider incorporating Bayesian model checking, such as
> posterior predictive checks.
>

True, Bayesians don't like to use chi^2 as a goodness-of-fit measure.  But in this case we're not really doing that.  We are instead using this to ask if the compression scheme has successfully reproduced the parts of the n(z) distribution of the samples that are actually relevant to predicting the c values.  If it has been successful, then the green and orange histograms (input samples vs samples from our generative model) will match.  So we think the plot is informative of the goal.

> 37. Figure 1: The input and generated samples align only near redshift
> 0.88. Could the authors explain this discrepancy at other redshifts?
>

Is the referee referring to the fact that the blue violins are shorter than the orange ones? The means of the two distributions (centers of the violins) are, by construction, identical.  The blue, compression-derived distributions are shorter (narrower distributions) than the orange input distributions, which is a result of projecting away modes of n that have no influence on the observable c values.  Removing modes can only reduce the variance of n(z) at a particular z, so the blue violins are shorter than the orange ones, by varying degrees at different values of z.

> 38. General: The method aims to sample q, but evidence of improved q sampling is
> limited. Could the authors provide empirical results demonstrating enhanced q
> sampling or discuss the scientific impact of improved n sampling? A numerical
> example showing how the method affects scientific conclusions compared to
> existing approaches would be valuable.
>

Our application is complex enough that it is difficult to come up with a single bottom-line measurement that says that our compression scheme is better than some other alternative for running a Markov chain over an 80-dimensional vector.  As noted in the introduction, it is simply not feasible to build a density estimator directly in the 80d space, nor to concatenate MC's that fix n to each sampled values, nor to use the hypercube discrete mapping - so no comparison of sampling results is even possible. We do demonstrate in Figure 4 that the ad-hoc resampling scheme for n(z) used in previous DES analyses (Eqn 27) does fail in at least one respect to reproduce the effect of the true n(z) samples on the cosmological observables, while the mode-projection method succeeds.  The difference is, admittedly, not dramatic, but at least we now have a method that is principled and fast.

We have inserted text into the concluding section to reiterate these points.

> I hope these suggestions are helpful in refining the manuscript. The proposed
> method has potential, and addressing these points could enhance its clarity and
> impact.
>

Thank you!

### 2025-08-22 main.tex
Resubmission version

@@
 % \documentclass[preprint,linenumbers]{aastex631} # 6.31 fails with \align!!
-%\documentclass[linenumbers, onecolumn]{aastex7}
-\documentclass[onecolumn]{aastex7}
+\documentclass[linenumbers, onecolumn, resetfootnote]{aastex7}
+%\documentclass[onecolumn]{aastex7}
 \usepackage{xcolor}
@@
 \newcommand{\vecv}{\ensuremath{\mathbf{v}}}
-\newcommand{\hatc}{\ensuremath{\hat{\mathbf{c}}}}
+\newcommand{\hatc}{\ensuremath{\bar{\mathbf{c}}}}
 \newcommand{\vecP}{\ensuremath{\mathbf{P}}}
@@
 
-\title{Sampling Bayesian probabilities given only sampled priors}
+\title{Sampling posterior distributions when only samples from a prior are available}
 
@@
 from the posterior $p(\vecq,\vecn | D) \propto \likeli(D | \vecq,\vecn)
-p(\vecq) p(\vecn),$ where $\vecn$ are some nuisance parameters.  In
-many cases, the nuisance parameters are high-dimensional, and their
+p(\vecq) p(\vecn),$ where $\vecn$ are some nuisance parameters \ed{with
+  separable prior.}  In \ed{some}
+cases, the nuisance parameters are high-dimensional, and their
 prior $p(\vecn)$ is itself defined only by a set of samples that have
@@
 interest \vecq.  There is a model $\hatc(\vecq,\vecn)$ for the
-observables which involves the parameters of interest, but also a
+\ed{expectation value of the} observables which involves the parameters of interest, but also a
 vector \vecn\ of nuisance parameters.  
-We wish to characterize the Bayesian posterior probability
+We wish to characterize the Bayesian posterior \ed{density}
 \begin{equation}
@@
 \end{equation}
-where $\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function
-of the data, and $p(\vecq)$ and $p(\vecn)$ are priors on the
-parameters.  This posterior is complex enough that it requires
+where \ed{$\likeli(\vecc | \vecq, \vecn)$ is a known likelihood function
+of the data, and we assume that the priors on $\vecq$ and $\vecn$ are separable
+to $p(\vecq)$ and $p(\vecn)$.}\footnote{Following the nomenclature for the
+elements of Bayes's formula in, \eg, \url{https://en.wikipedia.org/wiki/Posterior_probability}.}
+This posterior is complex enough that it requires
 approximation by the output of a Markov Chain (MC) wandering across the space $(\vecq,\vecn).$
+\ed{Implicit in \eqq{eq:posterior} is that the prior $p(\vecn)$ is independent of
+  the likelihood $\likeli$ for $\vecc,$ \eg\ the prior on $\vecn$ has been constrained using
+  data that are distinct from those entering the likelihood.}
 
-The scenario we address here is when \emph{the prior $p(\vecn)$ is
-  itself known only from a set of samples of $\vecn$ from this
-  distribution.} Most MC samplers require that the posterior (and
+The scenario we address here is when \emph{the prior $p(\vecn)$ is \ed{not
+    available in evaluable form, rather we have only
+  a set of samples of $\vecn$ known to be drawn} from this
+  distribution,} Most MC samplers require that the posterior (and
 hence the prior and likelihood) be an
@@
 The $b_k$ are a set of predetermined basis functions for the redshift
-distribution.
-In our case of analyzing the data from the  \textit{Dark Energy Survey} (DES), there are 10 distinct samples of galaxies---each designed to prefer galaxies in a limited range of redshift---which we can index by $s.$ Each has its own $n_s(z)$ to be characterized by coefficients $n_{sk}$ at $\approx100$ values of $k$, leading to $N=O(1000)$ parameters $n_{sk}$ to be considered.
+distribution, \ed{\eg\ these are boxcar functions if we are modeling $n(z)$ as
+  stepwise constant.}  In our case of analyzing the data from the  \textit{Dark Energy Survey} (DES), there are 10 distinct samples of galaxies---each designed to prefer galaxies in a limited range of redshift---which we can index by $s.$ Each has its own $n_s(z)$ to be characterized by coefficients $n_{sk}$ at $\approx100$ values of $k$, leading to $N=O(1000)$ parameters $n_{sk}$ to be considered.
 The vector $\vecn$ of nuisance parameters would be the concatenation of all the $n_{sk}.$ For clarity, we will still write this as $\vecn=\{n_1,n_2,\ldots,n_N\}$ and demonstrate the method with a single galaxy sample's $n(z)$.
 
-One approach would be to run a new MC chain over \vecq\ for each of the samples we have of \vecn, and then concatenate these to effect marginalization over \vecn.
-This is clearly infeasible if a large number of \vecn\ samples are needed to characterize the prior in this space.
+One approach would be to run a new MC over \vecq\ for each of the samples
+we have of \vecn, and then concatenate these to effect marginalization over
+\vecn.  This is clearly infeasible if a \ed{thousands or more} of \vecn\ samples are needed to characterize the prior in this space.
 
@@
 \citet{hyperrank} devised a scheme whereby
-the samples of \vecn\ are placed in a grid within some $M$-dimensional
+the samples of \vecn, \ed{which we write as $\{\vecn_\alpha\}$ for $\alpha\in
+  1\ldots N_\alpha,$} are placed in a grid within some $M$-dimensional 
 unit hypercube $\mathcal{H}$.  The coordinates \vecu\ within the hypercube
@@
   \vecu.  This solves the problem of creating a continuous \vecu\
-  domain, and maintains the equal prior probability of each \vecn\
-  sample, but the function \emph{output,} and the resultant likelihood
-  function of \vecu, are discontinuous.
+  domain, and \ed{gives each sample $\vecn_\alpha$ equal probability under
+  $p(\vecu),$ maintaining the meaning of the input samples.}
+But the \ed{\emph{output} of the function $\hat\vecn(\vecu)$,} and the resultant likelihood
+  function \ed{$\likeli(\vecc | \vecq, \vecu),$ are discontinuous over $\vecu.$}
 Various strategies are proposed by \citet{hyperrank} to assign the $\vecn_\alpha$ to the grid points in
@@
 \end{enumerate}
-Under these conditions, the marginalization over \vecn\ is equivalent to adding terms to $\covm_c,$ such that any MC process need not sample \vecn\ at all.
+Under these conditions, the marginalization over \vecn\ is \ed{shown to be
+  algebraically} equivalent to adding terms to $\covm_c,$ such that any MC process need not sample \vecn\ at all.
 
@@
 but does not require the 2nd condition of Gaussianity for
-the nuisance prior, and is likely to be a good approximation when
-the first and third conditions are not grossly violated.
-Our approach is to seek a linear compression of \vecn\ into a lower-dimensional set of parameters \vecu\ that projects away variations in \vecn\ that do not influence the likelihood $\likeli.$  Standard density estimators can then be applied to the \vecu\ values implied by the known \vecn\ samples to yield a prior $p(\vecu)$ that can be used for the MC chain of the cosmological posterior.
-The model \hatc, and hence $\likeli,$ will be continuous over this low-dimensional \vecu\ space, and marginalization over \vecu\ will yield posterior probabilities very close to marginalization over the original \vecn.
+the nuisance prior.
+%, and is likely to be a good approximation when the first and third conditions
+%are not grossly violated.
+Our approach is to seek a linear compression of \vecn\ into a lower-dimensional
+set of parameters \vecu\ that projects away variations in \vecn\ that do not
+influence the likelihood $\likeli.$  Standard density estimators can then be
+applied to the \vecu\ values implied by the known \vecn\ samples to yield a
+prior $p(\vecu)$ that can be used for the MC of the cosmological posterior.  The
+model \hatc, and hence $\likeli,$ will be continuous over this low-dimensional
+\vecu\ space, and marginalization over \vecu\ will yield posterior probabilities
+very close to marginalization over the original \vecn.  \ed{In the example
+  application described in Section~\ref{sec:app}, an 80-dimensional \vecn\ is
+  compressed into a 3-dimensional space.}
 
@@
 observables \vecc\, with the mean being some model
-$\hatc(\vecq,\vecn)$ and a fixed covariance matrix $\covm_c.$ In this
+$\hatc(\vecq,\vecn)$ and a fixed, \ed{known} covariance matrix $\covm_c.$ In this
 section we will assume that the \vecn\ vectors have been shifted by
-the mean of the samples $\vecn_0 \equiv \left\langle \vecn_\alpha \right\rangle$ so
-that the new \vecn\ has zero mean.
+the mean of the samples $\vecn_0 \equiv \left\langle \vecn_\alpha
+\right\rangle,$ \ed{generating a replacement set of $n_\alpha$ that have zero mean.}
 We seek some function $\hat{\vecn}(\vecu)$ of a much lower-dimensional
@@
 same likelihood function for any \vecn\ in the domain spanned by the
-samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha.$
+samples $\{\vecn_\alpha\}.$
 This means we want maps $\vecn_\alpha\rightarrow
@@
 \eqq{eq:posterior}---rather, $\left\langle\chi^2\right\rangle$ is serving as a
-proxy for the true log-likelihood shift.
+proxy for the true log-likelihood shift.  \ed{It is true, at least, that as $\langle
+\chi^2 \rangle \rightarrow 0,$ that the substitution of $\hat\vecn$ for $\vecn$
+has no effect on the calculated posterior density, so can be considered a
+lossless compression as far as inference on \vecq\ is concerned.}
 
@@
     minimum that keeps the $\left\langle\chi^2\right\rangle$ value in
-    \eqq{eq:chiresid} below a chosen threshold, presumably $\ll 1.$
+    \eqq{eq:chiresid} below a chosen threshold, presumably $\ll 1.$ \ed{If this
+      is achieved, the multiplicative errors in the likelihood induced by the
+      compression are at percent
+      levels, so errors in the posterior likelihoods are bounded to the same level.}
   \item The encoding matrix $E$ and decoding matrix $D$ are formed as
@@
 particular redshift range, as described in \citet{y6maglim}.  Once each bin's
-sample is chosen, a combination of 
+\ed{galaxy selection criteria are} chosen, a combination of 
 photometric techniques \citep{y6lenspz,y6pz} and clustering information
 \citep{y6wz} is used to generate samples from the posterior
-distribution of  the $n(z)$ distribution of the bin members, conditioned
-on the photometric and clustering data.
+distribution of  \ed{$n(z)$ functions applicable to the bin members,} conditioned
+on the photometric and clustering data, \ed{\ie\ the $p(\vecn)$ that will become the
+  prior for the inference of \vecq.}
 
@@
 dimensionality to those directions in $\vecn$ space in which the samples span a large
-enough range to alter the $\hatc(\vecn)$ at detectable levels.
+enough range to alter the $\hatc(\vecn)$ at \ed{levels comparable to the
+  precision of the observations.}
 
@@
 \caption{Violin plots for the redshift probability distribution $n(z)$
-  of galaxies in lens bin 4 for the DES Y6 analysis.  The blue regions
+  of galaxies in lens bin 4 for the DES Y6 analysis.  The orange regions
   show the distributions for the samples of $\vecn$ derived from
-  photometric and clustering information.  The orange violins are for
+  photometric and clustering information.  The blue violins are for
   $\vecn$ values drawn defined by (1) subtracting the mean $\bar\vecn$; (2) compressing these $\vecn$
@@
 \eqq{eq:chiresid}.  This modelling error induced by compression drops
-exponentially with the number of retained modes.
-We have somewhat arbitrarily chosen a threshold of $\chi^2<0.025$ for each galaxy sample in order to keep the total impact from modal compression $\ll 1.$ This is attained with $M=3$ modes for this bin.
-The right side of Figure~\ref{fig:chiresid} plots the individual modes
+exponentially with the number of retained modes.  We have somewhat arbitrarily
+chosen a threshold of $\chi^2<0.025$ for each galaxy sample in order to keep the
+total impact of compression \ed{of 10 samples} to $\ll 1.$ This is attained with $M=3$ modes for this bin.  The right side of Figure~\ref{fig:chiresid} plots the individual modes
 ${\rm U}_i(z),$ \ie\ the rows of the decompression matrix $\matD$ such
@@
 
-The orange violins in Figure~\ref{fig:violins} show the distributions
+The blue violins in Figure~\ref{fig:violins} show the distributions
 of the $n(z)$ values that are generated by drawing $\widetilde{\vecu}$
@@
   aim of making it feasible to produce continuous density estimators
-  for nuisance parameters that are known only from a set of samples in
+  for nuisance parameters that \ed{have no evaluable prior, and} are characterized only
+  from a set of samples in 
   a high-dimensional space $\vecn.$  This is essentially a principle
@@
 
-  We have shown the utility of this technique for the case when
-  $\vecn$ represents the redshift distributions of galaxies in DES.
-  The resulting modeled distribution of $n(z)$'s is more faithful to
-  the original sample than the previous \textit{ad hoc} method of
-  shifting and stretching the mean $n(z)$ to mimic the samples'
-  range.  While this difference has minimal consequence for the
+\ed{This technique was developed} for the case when
+  $\vecn$ represents the redshift distributions of galaxies in DES.  \ed{Some
+  attempts at sampling the posterior $p(\vecq| \vecc)$ while fully respecting
+  the prior $p(\vecn)$ as embodied by the samples $\{\vecn_\alpha\}$ have
+  proven infeasible, such as concatenating MC's run at every $\vecn_\alpha,$ or the
+  nearest-neighbor hypercube embedding of \citet{hyperrank}.  The method of
+  \citet{bridle02} that propagates $p(\vecn)$ into an inflated $\covm_c$ relies on
+  assumptions of Gaussianity for the nuisances that are not necessarily valid
+  here, and precludes some forms of internal consistency checks.  This forced
+  previous DES analyses to create an \textit{ad hoc} family of parameterized
+  $n(z)$ functions and priors in an attempt to mimic the cosmological impact of
+  the true $p(\vecn)$ distribution.  The mode-projection technique we describe is
+  able to do a better job, with
+  principled adherence to the prior embodied by
+  the input samples.
+  Extensive
+  use of this method on the 10 different galaxy samples defined for the Year 6
+  analysis of DES data is reported by \citet{y6pz, y6maglim} and \citet{y6wz}.
+
+While this new approach to marginalizing over $n(z)$  has minimal consequence for the
   posterior cosmological parameter estimates in the DES Y6 analysis,
@@
   accurate posteriors in future experiments where the $n(z)$
-  prior's uncertainty is a dominant contributor to the cosmological posterior.
+  prior's uncertainty is a dominant contributor to the cosmological posterior,
+  and can be generally applicable to other analyses where critical nuisance parameters
+  do not have evaluable priors.}
 
